---
title: "Species distribution per site"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load(tidyverse,
               vegan)
```

# Load data

```{r}
A.1_Metadata <- 
  readxl::read_xlsx("metadata_20251021.xlsx", sheet = "Sheet1")
```

# Bars

Tidy.

```{r}
A.1_Metadata_1 <- 
  A.1_Metadata |> 
  select(2, 3, 12)
```

Count.

```{r}
A.1_Metadata_2 <- 
  A.1_Metadata_1 |> 
  group_by(region) |> 
  count(Species)
```

Plot.

```{r}
A.2_Metadata_2 <- 
  A.1_Metadata_2 |> 
  ggplot(aes(y = region, x = n, fill = Species)) +
  geom_col() +
  theme_classic()
```

# Clustering

Tidy.

```{r}
B.1_Metadata <- 
  A.1_Metadata_2 |> 
  ungroup() |> 
  pivot_wider(names_from = Species,
              values_from = n,
              values_fill = 0) 

B.1_Matrix <- 
  B.1_Metadata |> 
  column_to_rownames("region")

B.1_Matrix_1 <- # Hellinger transform
  B.1_Matrix |> 
  decostand(method = "hellinger")

B.1_Matrix_2 <- # BC
  B.1_Matrix_1 |> 
  vegdist(method = "bray")

B.tre_WardD2 <-
  B.1_Matrix_2 |> 
  hclust(method = "ward.D2")

B.vec_RegionOrder <- 
  B.tre_WardD2$labels[B.tre_WardD2$order]
```

# Bars 2

```{r}
C.2_Metadata <- 
  A.1_Metadata_2 |> 
  ggplot(aes(y = fct_relevel(region, B.vec_RegionOrder), x = n, fill = Species)) +
  geom_col() +
  theme_classic()
```

# Bars 3

Load ITS2 data.

```{r}
C.2_ITS2 <- 
  read_tsv('../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt')
```

Tidy.

```{r}
C.2_ITS2_1 <- 
  C.2_ITS2 |> 
  filter(sample_uid != "seq_accession")

C.2_ITS2_2 <- # correct ids
  C.2_ITS2_1 |> 
  left_join((A.1_Metadata |> select(1,2) |> 
               rename(sample_name = 1))) |> 
  select(!sample_name) |> 
  relocate(CorrectID)

C.2_ITS2_3 <- # filter samples with ITS2 seqs
  C.2_ITS2_2 |> 
  filter(post_med_absolute != 0)

C.vec_ITS2 <- 
  C.2_ITS2_3 |> 
  pull(CorrectID)
```

Recluster.

```{r}
D.1_Metadata_1 <- 
  A.1_Metadata_1 |> 
  filter(CorrectID %in% C.vec_ITS2) |> 
  group_by(region) |> 
  count(Species) 

D.1_Metadata_2 <- 
  D.1_Metadata_1 |>
  ungroup() |> 
  pivot_wider(names_from = Species,
              values_from = n,
              values_fill = 0) 

D.1_Matrix <- 
  D.1_Metadata_2 |> 
  column_to_rownames("region")

D.1_Matrix_1 <- # Hellinger transform
  D.1_Matrix |> 
  decostand(method = "hellinger")

D.1_Matrix_2 <- # D.
  D.1_Matrix_1 |> 
  vegdist(method = "bray")

D.tre_WardD2 <-
  D.1_Matrix_2 |> 
  hclust(method = "ward.D2")

D.vec_RegionOrder <- 
  D.tre_WardD2$labels[D.tre_WardD2$order]
```

Reviz.

```{r}
D.vec_Species <- 
  c("H. hippopus",
    "H. porcellanus",
    "T. crocea",
    "T. squamosa",
    "T. noae",
    "T. maxima",
    "T. derasa",
    "T. gigas")

D.vec_SpeciesColors <- 
  c("H. hippopus" = "darkred",
    "H. porcellanus" = "#cf202e",
    "T. crocea" = "#f78822",
    "T. squamosa" = "#f6db35",
    "T. noae" = "#6dc067",
    "T. maxima" = "#4185be",
    "T. derasa" = "#8f499c",
    "T. gigas" = "#201146")

D.2_Metadata <- 
  D.1_Metadata_1 |> 
  ggplot(aes(y = fct_relevel(region, D.vec_RegionOrder),
             x = n, 
             fill = fct_relevel(Species, D.vec_Species))) +
  scale_fill_manual(values = D.vec_SpeciesColors) +
  geom_col() +
  scale_x_continuous(expand = c(0,0),
                     limits = c(0,25)) +
  theme_classic() +
  theme(legend.text = element_text(face = "italic"),
        axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(x = "Number of colonies",
       y = "Sampling region",
       fill = "Giant clam species")

# Save with legend
D.2_Metadata |> 
  ggsave(filename = "Species distribution.svg", 
         dpi = 900,
         width = 8,
         height = 6, 
         units = "cm")

# Save without legend
(D.2_Metadata +
  theme(legend.position = "none")) |> 
  ggsave(filename = "Species distribution_1.svg", 
         dpi = 900,
         width = 5.5,
         height = 6, 
         units = "cm")
```
