---
title: "Indicator species analysis"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load("tidyverse",
               "indicspecies",
               "vegan",
               "compositions",
               "janitor")
```

# Genus

Load data.

```{r}
# ITS2 data
A.1_ITS2 <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

A.1_ITS2_1 <- 
  A.1_ITS2 |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

# Metadata
A.1_Metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx")

A.1_Metadata_1 <- # rename col ID
  A.1_Metadata  |> 
  select(c(1, 2, 3, 12)) |> 
  rename(sample_name = 1)

A.1_ITS2_2 <- 
  A.1_ITS2_1 |> 
  left_join(A.1_Metadata_1) |> 
  relocate(sample_name, CorrectID, Species, region)

A.1_Metadata_2 <- # filter metadata
  A.1_Metadata_1 |>
  filter(CorrectID %in% (A.1_ITS2_2 |> pull(CorrectID)))
```

Get genus counts.

```{r}
B.1_Genus <- 
  A.1_ITS2_2 |> 
  select(!c(1,3,4)) |> 
  pivot_longer(!1,
               names_to = "ITS2",
               values_to = "Count")  |> 
  left_join(A.1_Metadata_2)

B.1_Genus_1 <- 
  B.1_Genus |> 
  # Detect genus
  mutate(Genus = case_when(str_detect(ITS2, "A") ~ "Symbiodinium",
                           str_detect(ITS2, "C") ~ "Cladocopium",
                           str_detect(ITS2, "D") ~ "Durusdinium",
                           str_detect(ITS2, "G") ~ "Gerakladium",
                           .default = NA)) |> 
  drop_na() |> 
  group_by(CorrectID, Genus, Species, region) |> 
  summarise(Total = sum(Count)) |> 
  ungroup()

B.1_Genus_2 <- # community data
  B.1_Genus_1 |> 
  pivot_wider(names_from = Genus,
              values_from = Total) |> 
  select(-c(2,3)) |> 
  arrange(CorrectID) |> 
  column_to_rownames("CorrectID") |> 
  as.matrix()

B.1_GenusSpecInfo <- # species info
  A.1_Metadata_2 |> 
  arrange(CorrectID) |> 
  pull(Species) |> 
  factor()

B.1_GenusRegInfo <- # species info
  A.1_Metadata_2 |> 
  arrange(CorrectID) |> 
  pull(region) |> 
  factor()
```

Run indicator species analysis.

```{r}
C.1_ISA <- 
  multipatt(B.1_Genus_2, B.1_GenusSpecInfo, 
                    control = how(nperm=999))

summary(C.1_ISA, indvalcomp = TRUE)

C.1_ISA_1 <- 
  multipatt(B.1_Genus_2, B.1_GenusSpecInfo, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(C.1_ISA_1, indvalcomp = TRUE)
```

Summary: not a good way to show dominance, stick with Friedman's test.

# ITS2 sequences

Tidy data. Do not filter.

```{r}
C.1_ITS2 <- 
  A.1_ITS2_2 |> 
  arrange(CorrectID) |> 
  select(-c(1,3,4)) |> 
  column_to_rownames("CorrectID")

C.1_GenusSpecInfo <- # species info
  A.1_Metadata_2 |> 
  arrange(CorrectID) |> 
  pull(Species) |> 
  factor()

C.1_GenusRegInfo <- # species info
  A.1_Metadata_2 |> 
  arrange(CorrectID) |> 
  pull(region) |> 
  factor()
```

Run ISA.

```{r}
D.1_ISA <- 
  multipatt(C.1_ITS2, C.1_GenusSpecInfo, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(D.1_ISA, indvalcomp = TRUE)

D.1_ISA_1 <- 
  multipatt(C.1_ITS2, C.1_GenusRegInfo, 
                    control = how(nperm=999),
            duleg = TRUE)


```

Filter to get the DIVs only.

```{r}
E.1_DIVs <- 
  read_tsv("../DIVs.txt",col_names = F) |> 
  pull()

E.1_DIVs_1 <- 
  A.1_ITS2_2 |> 
  pivot_longer(-c(1:4),
               names_to = "ITS2",
               values_to = "Count") |> 
  filter(ITS2 %in% E.1_DIVs) |> 
  pivot_wider(names_from = "ITS2",
              values_from = "Count") |> 
  arrange(CorrectID) |> 
  select(-c(1,3,4)) |> 
  column_to_rownames("CorrectID")

E.1_GenusSpecInfo <- # species info
  A.1_Metadata_2 |> 
  arrange(CorrectID) |> 
  pull(Species) |> 
  factor()

E.1_GenusRegInfo <- # species info
  A.1_Metadata_2 |> 
  arrange(CorrectID) |> 
  pull(region) |> 
  factor()

# Run ISA
E.1_ISA <- 
  multipatt(E.1_DIVs_1, C.1_GenusSpecInfo, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(E.1_ISA, indvalcomp = TRUE)


E.1_ISA_1 <- 
  multipatt(E.1_DIVs_1, E.1_GenusRegInfo, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(E.1_ISA_1, indvalcomp = TRUE)
```

# ITS2 type profiles

Load.

```{r}
F.1_ITS2TypeProf <- 
  read_tsv("../raw_data/20250829T091611/its2_type_profiles/4_first_analysis_20250829T091611.profiles.absolute.abund_and_meta.txt")

F.1_ITS2TypeProf_1 <- # tidy
  F.1_ITS2TypeProf |> 
  slice(-(1:5)) %>% # Remove first 5 rows (unnecessary rows)
  row_to_names(1) |>  # Move first row to column header
  rename(sample_name = 2) |>  # Name 2nd column
  select(-1) |>  # Remove first column (unneeded)
  drop_na() |> 
  type_convert()

F.1_ITS2TypeProf_2 <- 
  A.1_Metadata_2 |> 
  left_join(F.1_ITS2TypeProf_1) |> 
  arrange(CorrectID) |> 
  select(-c(1,3,4)) |> 
  column_to_rownames("CorrectID")
```

ISA.

```{r}
F.1_ISA <- 
  multipatt(F.1_ITS2TypeProf_2, C.1_GenusSpecInfo, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(F.1_ISA, indvalcomp = TRUE)

F.1_ISA_1 <- 
  multipatt(F.1_ITS2TypeProf_2, C.1_GenusRegInfo, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(F.1_ISA_1, indvalcomp = TRUE)
```

# Comparison within species

*H. hippopus*.

```{r}
G.1_DIVsHhip <- 
  A.1_ITS2_2 |> 
  pivot_longer(-c(1:4),
               names_to = "ITS2",
               values_to = "Count") |> 
  filter(ITS2 %in% E.1_DIVs) |> 
  pivot_wider(names_from = "ITS2",
              values_from = "Count") |> 
  filter(Species == "H. hippopus")

G.1_DIVsHhip_1 <- 
  G.1_DIVsHhip |> 
  arrange(CorrectID) |> 
  select(-c(1,3,4)) |> 
  column_to_rownames("CorrectID")

# ISA
G.1_GenusRegInfoHhip <- # species info
  G.1_DIVsHhip |> 
  arrange(CorrectID) |> 
  pull(region) |> 
  factor()

G.1_ISAHhip <- 
  multipatt(G.1_DIVsHhip_1, G.1_GenusRegInfoHhip, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(G.1_ISAHhip, indvalcomp = TRUE)
```

*H. porcellanus.*

```{r}
G.1_DIVsHpor <- 
  A.1_ITS2_2 |> 
  pivot_longer(-c(1:4),
               names_to = "ITS2",
               values_to = "Count") |> 
  filter(ITS2 %in% E.1_DIVs) |> 
  pivot_wider(names_from = "ITS2",
              values_from = "Count") |> 
  filter(Species == "H. porcellanus")

G.1_DIVsHpor_1 <- 
  G.1_DIVsHpor |> 
  arrange(CorrectID) |> 
  select(-c(1,3,4)) |> 
  column_to_rownames("CorrectID")

# ISA
G.1_GenusRegInfoHpor <- # species info
  G.1_DIVsHpor |> 
  arrange(CorrectID) |> 
  pull(region) |> 
  factor()

G.1_ISAHpor <- 
  multipatt(G.1_DIVsHpor_1, G.1_GenusRegInfoHpor, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(G.1_ISAHpor, indvalcomp = TRUE)
```

*T. crocea.*

```{r}
G.1_DIVsTcro <- 
  A.1_ITS2_2 |> 
  pivot_longer(-c(1:4),
               names_to = "ITS2",
               values_to = "Count") |> 
  filter(ITS2 %in% E.1_DIVs) |> 
  pivot_wider(names_from = "ITS2",
              values_from = "Count") |> 
  filter(Species == "T. crocea")

G.1_DIVsTcro_1 <- 
  G.1_DIVsTcro |> 
  arrange(CorrectID) |> 
  select(-c(1,3,4)) |> 
  column_to_rownames("CorrectID")

# ISA
G.1_GenusRegInfoTcro <- # species info
  G.1_DIVsTcro |> 
  arrange(CorrectID) |> 
  pull(region) |> 
  factor()

G.1_ISATcro <- 
  multipatt(G.1_DIVsTcro_1, G.1_GenusRegInfoTcro, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(G.1_ISATcro, indvalcomp = TRUE)
```

*T. squamosa.*

```{r}
G.1_DIVsTsqu <- 
  A.1_ITS2_2 |> 
  pivot_longer(-c(1:4),
               names_to = "ITS2",
               values_to = "Count") |> 
  filter(ITS2 %in% E.1_DIVs) |> 
  pivot_wider(names_from = "ITS2",
              values_from = "Count") |> 
  filter(Species == "T. squamosa")

G.1_DIVsTsqu_1 <- 
  G.1_DIVsTsqu |> 
  arrange(CorrectID) |> 
  select(-c(1,3,4)) |> 
  column_to_rownames("CorrectID")

# ISA
G.1_GenusRegInfoTsqu <- # species info
  G.1_DIVsTsqu |> 
  arrange(CorrectID) |> 
  pull(region) |> 
  factor()

G.1_ISATsqu <- 
  multipatt(G.1_DIVsTsqu_1, G.1_GenusRegInfoTsqu, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(G.1_ISATsqu, indvalcomp = TRUE)
```

*T. noae.*

```{r}
G.1_DIVsTnoa <- 
  A.1_ITS2_2 |> 
  pivot_longer(-c(1:4),
               names_to = "ITS2",
               values_to = "Count") |> 
  filter(ITS2 %in% E.1_DIVs) |> 
  pivot_wider(names_from = "ITS2",
              values_from = "Count") |> 
  filter(Species == "T. noae")

G.1_DIVsTnoa_1 <- 
  G.1_DIVsTnoa |> 
  arrange(CorrectID) |> 
  select(-c(1,3,4)) |> 
  column_to_rownames("CorrectID")

# ISA
G.1_GenusRegInfoTnoa <- # species info
  G.1_DIVsTnoa |> 
  arrange(CorrectID) |> 
  pull(region) |> 
  factor()

G.1_ISATnoa <- 
  multipatt(G.1_DIVsTnoa_1, G.1_GenusRegInfoTnoa, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(G.1_ISATnoa, indvalcomp = TRUE)
```

*T. maxima*.

```{r}
G.1_DIVsTmax <- 
  A.1_ITS2_2 |> 
  pivot_longer(-c(1:4),
               names_to = "ITS2",
               values_to = "Count") |> 
  filter(ITS2 %in% E.1_DIVs) |> 
  pivot_wider(names_from = "ITS2",
              values_from = "Count") |> 
  filter(Species == "T. maxima")

G.1_DIVsTmax_1 <- 
  G.1_DIVsTmax |> 
  arrange(CorrectID) |> 
  select(-c(1,3,4)) |> 
  column_to_rownames("CorrectID")

# ISA
G.1_GenusRegInfoTmax <- # species info
  G.1_DIVsTmax |> 
  arrange(CorrectID) |> 
  pull(region) |> 
  factor()

G.1_ISATmax <- 
  multipatt(G.1_DIVsTmax_1, G.1_GenusRegInfoTmax, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(G.1_ISATmax, indvalcomp = TRUE)
```

*T. derasa.*

```{r}
G.1_DIVsTder <- 
  A.1_ITS2_2 |> 
  pivot_longer(-c(1:4),
               names_to = "ITS2",
               values_to = "Count") |> 
  filter(ITS2 %in% E.1_DIVs) |> 
  pivot_wider(names_from = "ITS2",
              values_from = "Count") |> 
  filter(Species == "T. derasa")

G.1_DIVsTder_1 <- 
  G.1_DIVsTder |> 
  arrange(CorrectID) |> 
  select(-c(1,3,4)) |> 
  column_to_rownames("CorrectID")

# ISA
G.1_GenusRegInfoTder <- # species info
  G.1_DIVsTder |> 
  arrange(CorrectID) |> 
  pull(region) |> 
  factor()

G.1_ISATder <- 
  multipatt(G.1_DIVsTder_1, G.1_GenusRegInfoTder, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(G.1_ISATder, indvalcomp = TRUE)
```

*T. gigas.*

```{r}
G.1_DIVsTgig <- 
  A.1_ITS2_2 |> 
  pivot_longer(-c(1:4),
               names_to = "ITS2",
               values_to = "Count") |> 
  filter(ITS2 %in% E.1_DIVs) |> 
  pivot_wider(names_from = "ITS2",
              values_from = "Count") |> 
  filter(Species == "T. gigas")

G.1_DIVsTgig_1 <- 
  G.1_DIVsTgig |> 
  arrange(CorrectID) |> 
  select(-c(1,3,4)) |> 
  column_to_rownames("CorrectID")

# ISA
G.1_GenusRegInfoTgig <- # species info
  G.1_DIVsTgig |> 
  arrange(CorrectID) |> 
  pull(region) |> 
  factor()

G.1_ISATgig <- 
  multipatt(G.1_DIVsTgig_1, G.1_GenusRegInfoTgig, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(G.1_ISATgig, indvalcomp = TRUE)
```

# T. gigas (source vs transloc.)

```{r}
H.1_Metadata <- #get Tgig only
  A.1_Metadata_2 |> 
  filter(Species == "T. gigas")
```
