---
title: "Community across biogeographic regions"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load(tidyverse,
               janitor,
               vegan,
               pairwiseAdonis,
               viridis)
```

# Community structure

Import metadata.

```{r}
A.1_Metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx")

A.1_MetadataClean <- 
  A.1_Metadata |> 
  select(SymportalID, CorrectID, Species, region) |> 
  clean_names()

A.1_MetadataBiogeo <- 
  read_csv("Biogeographic regions.csv")

A.1_MetadataCleanMerged <- 
  A.1_MetadataClean |> 
  left_join(A.1_MetadataBiogeo)
```

Load community data.

```{r}
A.1_ITS2 <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

A.1_MetadataCleanMerged2 <- 
  A.1_MetadataCleanMerged |> 
  rename(sample_name = 1) |> 
  arrange(correct_id)

A.1_ITS2_2 <- 
  A.1_ITS2 |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

A.1_ITS2_3 <- 
  A.1_ITS2_2 |> 
  left_join(A.1_MetadataCleanMerged2) |> 
  relocate(sample_name, correct_id, species, region, bio_region)
```

Make matrix.

```{r}
A.1_ITS2_4 <- # retain CorrectID only
  A.1_ITS2_3 |> 
  select(!c(1,3,4,5)) |> 
  arrange(correct_id)  |> 
  column_to_rownames("correct_id") |> 
  as.matrix() |> 
  sqrt()
```

Compute Bray-Curtis dissimilarity.

```{r}
A.1_ITS2BrayCurtis <- 
  A.1_ITS2_4 |> 
  vegdist()

A.2_PERMANOVA <- 
  adonis2(A.1_ITS2BrayCurtis ~ get("bio_region"), 
          data = (A.1_MetadataCleanMerged2 |> 
                    filter(correct_id %in% (A.1_ITS2_3 |>
                                             pull(correct_id))) |>
                    arrange(correct_id)))

A.2_PERMANOVAPairwise <- 
  pairwise.adonis2(A.1_ITS2BrayCurtis ~ bio_region,
                   data = (A.1_MetadataCleanMerged2 |>
                             filter(correct_id %in% (A.1_ITS2_3 |>
                                                      pull(correct_id))) |>
                      arrange(correct_id)),
                   nperm = 9999)

Z.fun_PPERMANOVACleaner <- 
  function(pmanova_res){
    a <- 
      pmanova_res[-1] |> 
      bind_rows(.id = 'groups') |> 
      drop_na() %>%
      .$`Pr(>F)` |> 
      p.adjust(method = 'BH') 
    
    b <- 
      unique(pmanova_res[-1] |> 
               bind_rows(.id = 'groups'))$groups |> 
      unique()
      
    d <- 
       a %>% 
        data.frame(b,.) |> 
        rename('groups' = 1,
               `P-adjusted` = 2) %>% 
        left_join(pmanova_res[-1] |> 
                    bind_rows(.id = 'groups'), 
                  ., 
                  by = 'groups') |> 
        mutate(`P-adjusted` = case_when(is.na(`Pr(>F)`) ~ NA,
                                        .default = `P-adjusted`)) |> 
        mutate(Parameter = rep(c('Species', 'Residual', 'Total'), length.out = n())) |> 
        relocate(last_col(), .after = 1)
    
    return(d)
  }

A.2_PERMANOVAPairwise2 <- # tidy
  A.2_PERMANOVAPairwise |> 
  Z.fun_PPERMANOVACleaner()
```

Plot.

```{r}
# Import ITS2 type profiles
B.1_ITS2TypeProf <- 
  read_tsv("../raw_data/20250829T091611/its2_type_profiles/4_first_analysis_20250829T091611.profiles.absolute.abund_and_meta.txt")

B.1_ITS2TypeProfClean <- 
  B.1_ITS2TypeProf |> 
  slice(-(1:5)) %>% # Remove first 5 rows (unnecessary rows)
  row_to_names(1) |>  # Move first row to column header
  rename(symportal_id = 2) |>  # Name 2nd column
  select(-1) |>  # Remove first column (unneeded)
  drop_na() |> 
  type_convert() 

B.1_ITS2TypeProfClean2 <- 
  B.1_ITS2TypeProfClean |> 
  left_join(A.1_MetadataClean) |> 
  left_join(A.1_MetadataBiogeo) |> 
  relocate(symportal_id, correct_id:bio_region) |> 
  rename(sample_id = 2)

# Correct ITS2 type profile naming
B.1_ITS2TypeProfCorrect <- 
  read_csv("../raw_data/Edited ITS2 type profile naming.csv") |> 
  rename(its2_type = 1) 

B.1_ITS2TypeProfCleanMerged <- 
  B.1_ITS2TypeProfClean2 |> 
  filter(sample_id %in% (A.1_ITS2_3 |> pull(correct_id))) |> 
  
  # Tidy
  pivot_longer(!c(symportal_id:bio_region),
               names_to = "its2_type",
               values_to = "count") |> 
  left_join(B.1_ITS2TypeProfCorrect)  |> 
  select(!its2_type) |> 
  rename(its2_type = Edited) |> 
  group_by(sample_id)  |>
  mutate(rel_abun = count/sum(count)) |> 
  ungroup() |> 
  mutate(rel_abun = sqrt(rel_abun),
         clade = str_extract(its2_type, "[A-Z]"))

B.2_ITS2TypeProfile <- 
  B.1_ITS2TypeProfCleanMerged |> 
  ggplot(aes(x = fct_reorder(sample_id, species),
             y = its2_type,
             fill = rel_abun)) +
  geom_raster() +
  facet_grid(cols = vars(bio_region),
             rows = vars(clade),
             space = "free",
             scale = "free") +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        strip.background = element_blank(),
        panel.spacing = unit(0.2,'lines'),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 7, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  viridis::scale_fill_viridis(option = "cividis")

# Export
B.2_ITS2TypeProfile |> 
  ggsave(filename = "By biogeographic regions.svg", 
         dpi = 900,
         width = 18, 
         height = 10,
         unit = "cm")

# Region labels
B.2_ITS2TypeProfile_region_lab <- 
  B.1_ITS2TypeProfCleanMerged |> 
  ggplot(aes(x = fct_reorder(sample_id,
                             species),
             y = 1,
             fill = species)) + 
  geom_raster() +
  facet_grid(cols = vars(bio_region),
             space = "free",
             scales = "free") +
  scale_fill_manual(values = 
                       c("H. hippopus"="darkred",
                         "H. porcellanus"="#cf202e",
                         "T. crocea"="#f78822",
                         "T. squamosa"="#f6db35",
                         "T. noae"="#6dc067",
                         "T. maxima"="#4185be",
                         "T. derasa"="#8f499c",
                         "T. gigas"="#201146")) +
  theme_void() +
  theme(legend.position = "bottom")

B.2_ITS2TypeProfile_region_lab |> 
  ggsave(filename = "By species - Labels.svg", 
         dpi = 900,
         width = 17.5, 
         height = 10,
         unit = "cm")
```

NMDS

```{r}
C.1_ITS2BrayCurtisNMDS <- 
  A.1_ITS2BrayCurtis |> 
  metaMDS(k = 2)

C.1_ITS2BrayCurtisNMDS_1 <- # tidy NMDS data
  as.data.frame(scores(C.1_ITS2BrayCurtisNMDS)) |> 
  rownames_to_column("correct_id")

C.1_ITS2BrayCurtisNMDS_2 <- # add metadata
  C.1_ITS2BrayCurtisNMDS_1 |> 
  left_join(A.1_MetadataCleanMerged2)

C.2_ITS2BrayCurtisNMDS <- 
  C.1_ITS2BrayCurtisNMDS_2 |> 
  ggplot(aes(x = NMDS1, y = NMDS2)) +
  geom_hline(yintercept = 0,
             color = 'darkgray',
             linetype = 'dashed',
             linewidth = 0.25) +
  geom_vline(xintercept = 0,
             color = 'darkgray',
             linetype = 'dashed',
             linewidth = 0.25)  +
  stat_ellipse(aes(group = bio_region,
                   color = bio_region),
               level = 0.75) +
  geom_point(aes(color = bio_region))+
  labs(x = "NMDS1",
       y = "NMDS2",
       color = "GC species") +
  scale_color_manual(values = c("West Philippine Sea"="#FDE725",
                               "Sulu Sea"="#7AD151",
                               "Visayas Sea"="#440154",
                               "North Philippine Sea"="#414487",
                               "South Philippine Sea"="#2A788E")) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.line = element_line(),panel.background    = element_rect(fill = "transparent", colour = NA),
        plot.background     = element_rect(fill = "transparent", colour = NA),
        legend.background   = element_rect(fill = "transparent", colour = NA),
        legend.box.background = element_rect(fill = "transparent", colour = NA),
        legend.position = "none")

C.2_ITS2BrayCurtisNMDS |> 
  ggsave(filename = "By biogeographic regions - NMDS.svg", 
         dpi = 900,
         width = 10, 
         height = 10,
         unit = "cm")
```
