---
title: "Mantel test"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load(tidyverse,
               vegan,
               janitor,
               ape,
               pairwiseAdonis)
```

# Mantel test

## *H. hippopus*

```{r}
metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

dna <- read.dna("Clean alignments per species/Hhippopus alignment - Clipped.fas",
                format="fasta")

host_dist <- 
  dist.dna(dna, model="K80")

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

community_cleaner <- 
  metadata |> 
  left_join(community_clean) |> 
  filter(sample_id %in% (dna |> rownames()))

sym_dist <- 
  community_cleaner |> 
  select(!1) |> 
  arrange(match(sample_id, (dna |> rownames()))) |> 
  column_to_rownames("sample_id") |> 
  vegdist(method="bray")

mantel_results <- 
  mantel(host_dist, sym_dist, method="spearman")

mantel_df <- tibble(
  host_distance = as.vector(host_dist),
  sym_distance  = as.vector(sym_dist)
)

mantel_plot <- 
  mantel_df |> 
  ggplot(aes(host_distance, sym_distance)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Host genetic distance (COI)",
    y = "Symbiodiniaceae community dissimilarity (Bray-Curtis)"
  ) +
  theme_classic()
```

## *H. porcellanus*

```{r}
metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

dna <- read.dna("Clean alignments per species/Hporcellanus alignment - Clipped.fas",
                format="fasta")

host_dist <- 
  dist.dna(dna, model="K80")

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

community_cleaner <- 
  metadata |> 
  left_join(community_clean) |> 
  filter(sample_id %in% (dna |> rownames()))

sym_dist <- 
  community_cleaner |> 
  select(!1) |> 
  arrange(match(sample_id, (dna |> rownames()))) |> 
  column_to_rownames("sample_id") |> 
  vegdist(method="bray")

mantel_results <- 
  mantel(host_dist, sym_dist, method="spearman")

mantel_df <- tibble(
  host_distance = as.vector(host_dist),
  sym_distance  = as.vector(sym_dist)
)

mantel_plot <- 
  mantel_df |> 
  ggplot(aes(host_distance, sym_distance)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Host genetic distance (COI)",
    y = "Symbiodiniaceae community dissimilarity (Bray-Curtis)"
  ) +
  theme_classic()
```

## *T. crocea*

```{r}
metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

dna <- read.dna("Clean alignments per species/Tcrocea alignment.fasta",
                format="fasta")

host_dist <- 
  dist.dna(dna, model="K80")

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

community_cleaner <- 
  metadata |> 
  left_join(community_clean) |> 
  filter(sample_id %in% (dna |> rownames()))

sym_dist <- 
  community_cleaner |> 
  select(!1) |> 
  arrange(match(sample_id, (dna |> rownames()))) |> 
  column_to_rownames("sample_id") |> 
  vegdist(method="bray")

mantel_results <- 
  mantel(host_dist, sym_dist, method="spearman")

mantel_df <- tibble(
  host_distance = as.vector(host_dist),
  sym_distance  = as.vector(sym_dist)
)

mantel_plot <- 
  mantel_df |> 
  ggplot(aes(host_distance, sym_distance)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Host genetic distance (COI)",
    y = "Symbiodiniaceae community dissimilarity (Bray-Curtis)"
  ) +
  theme_classic()
```

## *T. derasa*

```{r}
metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

dna <- read.dna("Clean alignments per species/Tderasa alignment - Clipped.fas",
                format="fasta")

host_dist <- 
  dist.dna(dna, model="K80")

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

community_cleaner <- 
  metadata |> 
  left_join(community_clean) |> 
  filter(sample_id %in% (dna |> rownames()))

sym_dist <- 
  community_cleaner |> 
  select(!1) |> 
  arrange(match(sample_id, (dna |> rownames()))) |> 
  column_to_rownames("sample_id") |> 
  vegdist(method="bray")

mantel_results <- 
  mantel(host_dist, sym_dist, method="spearman")

mantel_df <- tibble(
  host_distance = as.vector(host_dist),
  sym_distance  = as.vector(sym_dist)
)

mantel_plot <- 
  mantel_df |> 
  ggplot(aes(host_distance, sym_distance)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Host genetic distance (COI)",
    y = "Symbiodiniaceae community dissimilarity (Bray-Curtis)"
  ) +
  theme_classic()
```

## *T. gigas*

```{r}
metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

  dna <- read.dna("Clean alignments per species/Tgigas alignment.fasta",
                format="fasta")

host_dist <- 
  dist.dna(dna, model="K80")

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

community_cleaner <- 
  metadata |> 
  left_join(community_clean) |> 
  filter(sample_id %in% (dna |> rownames()))

sym_dist <- 
  community_cleaner |> 
  select(!1) |> 
  arrange(match(sample_id, (dna |> rownames()))) |> 
  column_to_rownames("sample_id") |> 
  vegdist(method="bray")

mantel_results <- 
  mantel(host_dist, sym_dist, method="spearman")

mantel_df <- tibble(
  host_distance = as.vector(host_dist),
  sym_distance  = as.vector(sym_dist)
)

mantel_plot <- 
  mantel_df |> 
  ggplot(aes(host_distance, sym_distance)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Host genetic distance (COI)",
    y = "Symbiodiniaceae community dissimilarity (Bray-Curtis)"
  ) +
  theme_classic()
```

## *T. maxima*

```{r}
metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

  dna <- read.dna("Clean alignments per species/Tmaxima alignment - Clipped.fas",
                format="fasta")

host_dist <- 
  dist.dna(dna, model="K80")

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

community_cleaner <- 
  metadata |> 
  left_join(community_clean) |> 
  filter(sample_id %in% (dna |> rownames()))

sym_dist <- 
  community_cleaner |> 
  select(!1) |> 
  arrange(match(sample_id, (dna |> rownames()))) |> 
  column_to_rownames("sample_id") |> 
  vegdist(method="bray")

mantel_results <- 
  mantel(host_dist, sym_dist, method="spearman")

mantel_df <- tibble(
  host_distance = as.vector(host_dist),
  sym_distance  = as.vector(sym_dist)
)

mantel_plot <- 
  mantel_df |> 
  ggplot(aes(host_distance, sym_distance)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Host genetic distance (COI)",
    y = "Symbiodiniaceae community dissimilarity (Bray-Curtis)"
  ) +
  theme_classic()
```

## *T. noae*

```{r}
metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

  dna <- read.dna("Clean alignments per species/Tnoae alignment.fasta",
                format="fasta")

host_dist <- 
  dist.dna(dna, model="K80")

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

community_cleaner <- 
  metadata |> 
  left_join(community_clean) |> 
  filter(sample_id %in% (dna |> rownames()))

sym_dist <- 
  community_cleaner |> 
  select(!1) |> 
  arrange(match(sample_id, (dna |> rownames()))) |> 
  column_to_rownames("sample_id") |> 
  vegdist(method="bray")

mantel_results <- 
  mantel(host_dist, sym_dist, method="spearman")

mantel_df <- tibble(
  host_distance = as.vector(host_dist),
  sym_distance  = as.vector(sym_dist)
)

mantel_plot <- 
  mantel_df |> 
  ggplot(aes(host_distance, sym_distance)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Host genetic distance (COI)",
    y = "Symbiodiniaceae community dissimilarity (Bray-Curtis)"
  ) +
  theme_classic()
```

## *T. squamosa*

```{r}
metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

  dna <- read.dna("Clean alignments per species/Tsquamosa alignment - Clipped.fas",
                format="fasta")

host_dist <- 
  dist.dna(dna, model="K80")

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

community_cleaner <- 
  metadata |> 
  left_join(community_clean) |> 
  filter(sample_id %in% (dna |> rownames()))

sym_dist <- 
  community_cleaner |> 
  select(!1) |> 
  arrange(match(sample_id, (dna |> rownames()))) |> 
  column_to_rownames("sample_id") |> 
  vegdist(method="bray")

mantel_results <- 
  mantel(host_dist, sym_dist, method="spearman")

mantel_df <- tibble(
  host_distance = as.vector(host_dist),
  sym_distance  = as.vector(sym_dist)
)

mantel_plot <- 
  mantel_df |> 
  ggplot(aes(host_distance, sym_distance)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Host genetic distance (COI)",
    y = "Symbiodiniaceae community dissimilarity (Bray-Curtis)"
  ) +
  theme_classic()
```

# PERMANOVA

## *H. hippopus*

```{r}
haplotype_meta <- 
  read_csv("Haplotype - Metadata.csv") |> 
  filter(str_detect(sample_id,pattern = "HH")) |> 
  mutate(haplotype = as.factor(haplotype)) |> 
  arrange(sample_id)

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

metadata_clean <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2) |> 
  filter(str_detect(sample_id, "HH"))

community_cleaner <- 
  metadata_clean |> 
  left_join(community_clean) |> 
  select(!1) |> 
  filter(sample_id %in% (haplotype_meta |> pull(sample_id))) |> 
  column_to_rownames("sample_id") |> 
  as.matrix()

beta <- 
  community_cleaner |> 
  vegdist(method="bray")

bc_permanova <- 
  adonis2(beta ~ haplotype,
          data = haplotype_meta,
          method = "bray",
          permutations = 999)
```

## *H. porcellanus*

```{r}
haplotype_meta <- 
  read_csv("Haplotype - Metadata.csv") |> 
  filter(str_detect(sample_id,pattern = "HP")) |> 
  mutate(haplotype = as.factor(haplotype)) |> 
  arrange(sample_id)

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

metadata_clean <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2) |> 
  filter(str_detect(sample_id, "HP"))

community_cleaner <- 
  metadata_clean |> 
  left_join(community_clean) |> 
  select(!1) |> 
  filter(sample_id %in% (haplotype_meta |> pull(sample_id))) |> 
  column_to_rownames("sample_id") |> 
  as.matrix()

beta <- 
  community_cleaner |> 
  vegdist(method="bray")

bc_permanova <- 
  adonis2(beta ~ haplotype,
          data = haplotype_meta,
          method = "bray",
          permutations = 999)
```

## *T. crocea*

```{r}
haplotype_meta <- 
  read_csv("Haplotype - Metadata.csv") |> 
  filter(str_detect(sample_id,pattern = "TC")) |> 
  mutate(haplotype = as.factor(haplotype)) |> 
  arrange(sample_id)

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

metadata_clean <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2) |> 
  filter(str_detect(sample_id, "TC"))

community_cleaner <- 
  metadata_clean |> 
  left_join(community_clean) |> 
  select(!1) |> 
  filter(sample_id %in% (haplotype_meta |> pull(sample_id))) |> 
  column_to_rownames("sample_id") |> 
  as.matrix()

beta <- 
  community_cleaner |> 
  vegdist(method="bray")

bc_permanova <- 
  adonis2(beta ~ haplotype,
          data = haplotype_meta,
          method = "bray",
          permutations = 999)
```

## *T. derasa*

```{r}
haplotype_meta <- 
  read_csv("Haplotype - Metadata.csv") |> 
  filter(str_detect(sample_id,pattern = "TD")) |> 
  mutate(haplotype = as.factor(haplotype)) |> 
  arrange(sample_id)

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

metadata_clean <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2) |> 
  filter(str_detect(sample_id, "TD"))

community_cleaner <- 
  metadata_clean |> 
  left_join(community_clean) |> 
  select(!1) |> 
  filter(sample_id %in% (haplotype_meta |> pull(sample_id))) |> 
  column_to_rownames("sample_id") |> 
  as.matrix()

beta <- 
  community_cleaner |> 
  vegdist(method="bray")

bc_permanova <- 
  adonis2(beta ~ haplotype,
          data = haplotype_meta,
          method = "bray",
          permutations = 999)
```

## *T. gigas*

```{r}
haplotype_meta <- 
  read_csv("Haplotype - Metadata.csv") |> 
  filter(str_detect(sample_id,pattern = "TG")) |> 
  mutate(haplotype = as.factor(haplotype)) |> 
  arrange(sample_id)

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

metadata_clean <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2) |> 
  filter(str_detect(sample_id, "TG"))

community_cleaner <- 
  metadata_clean |> 
  left_join(community_clean) |> 
  select(!1) |> 
  filter(sample_id %in% (haplotype_meta |> pull(sample_id))) |> 
  column_to_rownames("sample_id") |> 
  as.matrix()

beta <- 
  community_cleaner |> 
  vegdist(method="bray")

bc_permanova <- 
  adonis2(beta ~ haplotype,
          data = haplotype_meta,
          method = "bray",
          permutations = 999)
```

## *T. maxima*

```{r}
haplotype_meta <- 
  read_csv("Haplotype - Metadata.csv") |> 
  filter(str_detect(sample_id,pattern = "TM")) |> 
  mutate(haplotype = as.factor(haplotype)) |> 
  arrange(sample_id)

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

metadata_clean <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2) |> 
  filter(str_detect(sample_id, "TM"))

community_cleaner <- 
  metadata_clean |> 
  left_join(community_clean) |> 
  select(!1) |> 
  filter(sample_id %in% (haplotype_meta |> pull(sample_id))) |> 
  column_to_rownames("sample_id") |> 
  as.matrix()

beta <- 
  community_cleaner |> 
  vegdist(method="bray")

bc_permanova <- 
  adonis2(beta ~ haplotype,
          data = haplotype_meta,
          method = "bray",
          permutations = 999)
```

## *T. noae*

```{r}
haplotype_meta <- 
  read_csv("Haplotype - Metadata.csv") |> 
  filter(str_detect(sample_id,pattern = "TN")) |> 
  mutate(haplotype = as.factor(haplotype)) |> 
  arrange(sample_id)

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

metadata_clean <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2) |> 
  filter(str_detect(sample_id, "TN"))

community_cleaner <- 
  metadata_clean |> 
  left_join(community_clean) |> 
  select(!1) |> 
  filter(sample_id %in% (haplotype_meta |> pull(sample_id))) |> 
  column_to_rownames("sample_id") |> 
  as.matrix()

beta <- 
  community_cleaner |> 
  vegdist(method="bray")

bc_permanova <- 
  adonis2(beta ~ haplotype,
          data = haplotype_meta,
          method = "bray",
          permutations = 999)
```

## *T. squamosa*

```{r}
haplotype_meta <- 
  read_csv("Haplotype - Metadata.csv") |> 
  filter(str_detect(sample_id,pattern = "TS")) |> 
  mutate(haplotype = as.factor(haplotype)) |> 
  arrange(sample_id)

community <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

community_clean <- 
  community |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2)

metadata_clean <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx") |> 
  select(SymportalID, CorrectID) |> 
  rename(sample_name = 1, sample_id = 2) |> 
  filter(str_detect(sample_id, "TS"))

community_cleaner <- 
  metadata_clean |> 
  left_join(community_clean) |> 
  select(!1) |> 
  filter(sample_id %in% (haplotype_meta |> pull(sample_id))) |> 
  column_to_rownames("sample_id") |> 
  as.matrix()

beta <- 
  community_cleaner |> 
  vegdist(method="bray")

bc_permanova <- 
  adonis2(beta ~ haplotype,
          data = haplotype_meta,
          method = "bray",
          permutations = 999)
```
