---
title: "Haplotype and photosymbionts"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load(tidyverse,
               vegan,
               janitor,
               ape,
               ggtree)
```

# Community structure

Load metadata.

```{r}
A.1_Metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx")

A.1_MetadataClean <- 
  A.1_Metadata |> 
  select(CorrectID,
         Species,
         region) |> 
  rename(sample_id = 1,
         species = Species)

A.1_MetadataHaplo <- 
  read_csv("Haplotype - Metadata.csv")

# Merge metadata
A.1_MetadataCompleteMerged<- 
  A.1_MetadataHaplo |> 
  left_join(A.1_MetadataClean)
```

Load community data.

```{r}
B.1_ITS2TypeProf <- 
  read_tsv("../raw_data/20250829T091611/its2_type_profiles/4_first_analysis_20250829T091611.profiles.absolute.abund_and_meta.txt")

B.1_ITS2TypeProfClean <- 
  B.1_ITS2TypeProf |> 
  slice(-(1:5)) %>% # Remove first 5 rows (unnecessary rows)
  row_to_names(1) |>  # Move first row to column header
  rename(sample_name = 2) |>  # Name 2nd column
  select(-1) |>  # Remove first column (unneeded)
  drop_na() |> 
  type_convert() 

B.1_ITS2TypeProfClean2 <- 
  B.1_ITS2TypeProfClean |> 
  rename(sample_id = 1)

# Correct ITS2 type profile naming
B.1_ITS2TypeProfCorrect <- 
  read_csv("../raw_data/Edited ITS2 type profile naming.csv") |> 
  rename(its2_type = 1) 


B.1_ITS2TypeProfCleanMerged <- 
  A.1_MetadataCompleteMerged |> 
  left_join(B.1_ITS2TypeProfClean2) |> 
  
  # Tidy
  pivot_longer(!c(sample_id:region),
               names_to = "its2_type",
               values_to = "count") |> 
  left_join(B.1_ITS2TypeProfCorrect)  |> 
  select(!its2_type) |> 
  rename(its2_type = Edited) |> 
  group_by(sample_id)  |>
  mutate(rel_abun = count/sum(count)) |> 
  ungroup() |> 
  mutate(rel_abun = sqrt(rel_abun),
         clade = str_extract(its2_type, "[A-Z]"))
```

Plot.

```{r}
B.2_ITS2TypeProfile <- 
  B.1_ITS2TypeProfCleanMerged |> 
  ggplot(aes(x = fct_reorder(sample_id, haplotype),
             y = its2_type,
             fill = rel_abun)) +
  geom_raster() +
  facet_grid(cols = vars(species),
             rows = vars(clade),
             space = "free",
             scale = "free") +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        strip.background = element_blank(),
        panel.spacing = unit(0.2,'lines'),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 7, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  viridis::scale_fill_viridis(option = "cividis")

# Haplotype labels
B.2_ITS2TypeProfileHaplos <- 
  A.1_MetadataCompleteMerged |> 
  ggplot(aes(x = fct_reorder(sample_id,
                             haplotype),
             y = 1,
             fill = haplotype)) + 
  geom_raster() +
  facet_grid(cols = vars(species),
             space = "free",
             scales = "free") +
  scale_fill_distiller(palette = "YlGn") +
  theme_void() +
  theme(legend.position = "bottom")

# Export
B.2_ITS2TypeProfileHaplos |> 
  ggsave(filename = "Haplotypes.svg", 
         dpi = 900,
         width = 17.5, 
         height = 10,
         unit = "cm")

# Region labels
B.2_ITS2TypeProfileRegions <- 
  A.1_MetadataCompleteMerged |> 
  ggplot(aes(x = fct_reorder(sample_id,
                             haplotype),
             y = 1,
             fill = region)) + 
  geom_raster() +
  facet_grid(cols = vars(species),
             space = "free",
             scales = "free") +
  scale_fill_manual(values = c("Alaminos"="#E41A1C",
                               "Antique"="#377EB8",
                               "ARNP"="#4DAF4A",
                               "Batangas"="#984EA3",
                               "Bohol"="#FF7F00",
                               "Bolinao"="#FFFF33",
                               "Calaguas"="#A65628",
                               "Camiguin"="#F781BF",
                               "Palawan"="#999999",
                               "Samar"="#66C2A5",
                               "Zambales"="#FC8D62")) +
  theme_void() +
  theme(legend.position = "bottom")

B.2_ITS2TypeProfileRegions |> 
  ggsave(filename = "Regions.svg", 
         dpi = 900,
         width = 17.5, 
         height = 10,
         unit = "cm")

# Biogeographic regions
A.1_MetadataBiogeo <- 
  read_csv("Biogeographic regions.csv")

B.2_ITS2TypeProfileBiogeo <- 
  A.1_MetadataCompleteMerged |> 
  left_join(A.1_MetadataBiogeo) |> 
  ggplot(aes(x = fct_reorder(sample_id,
                             haplotype),
             y = 1,
             fill = bio_region)) + 
  geom_raster() +
  facet_grid(cols = vars(species),
             space = "free",
             scales = "free") +
  scale_fill_manual(values = c("West Philippine Sea"="#000000",
                               "Sulu Sea"="#E69F03",
                               "Visayas Sea"="#56B4E9",
                               "North Philippine Sea"="#139E73",
                               "South Philippine Sea"="#F0E441")) +
  theme_void() +
  theme(legend.position = "bottom")

B.2_ITS2TypeProfileBiogeo |> 
  ggsave(filename = "Biogeo.svg", 
         dpi = 900,
         width = 17.5, 
         height = 10,
         unit = "cm")
```

Re-arrange, add UniFrac data.

```{r}
# Get pre-made ITS2 type profile arrangements based on Unifrac data.
C.1_Cladocopium <- 
  read_csv("Cladocopium UniFrac - Order.csv",
           col_names = F) |> 
  pull(1)

C.1_Durusdinium <- 
  read_csv("Durusdinium UniFrac - Order.csv",
           col_names = F) |> 
  pull(1)

C.1_Symbiodinium <- 
  read_csv("Symbiodinium UniFrac - Order.csv",
           col_names = F) |> 
  pull(1)

C.1_AllUniFrac <- 
  c(rev(C.1_Symbiodinium), rev(C.1_Cladocopium), rev(C.1_Durusdinium))

# Replot
C.2_ITS2TypeProfile <- 
  B.1_ITS2TypeProfCleanMerged |> 
  ggplot(aes(x = fct_reorder(sample_id, haplotype),
             y = fct_relevel(its2_type, C.1_AllUniFrac),
             fill = rel_abun)) +
  geom_raster() +
  facet_grid(cols = vars(species),
             rows = vars(clade),
             space = "free",
             scale = "free") +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        strip.background = element_blank(),
        panel.spacing = unit(0.2,'lines'),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 7, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  viridis::scale_fill_viridis(option = "cividis")

# Export
C.2_ITS2TypeProfile |> 
  ggsave(filename = "ITS2 type profile - By haplotype.svg", 
         dpi = 900,
         width = 17.5, 
         height = 11,
         unit = "cm")
```
