---
title: "Haplotype"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load(ggplot2,
               dplyr,
               readr,
               tidyr,
               stringr,
               tibble,
               forcats,
               pegas,
               ape)
```

# Load data

```{r}
loc <- sample(c("north","south"),26,replace=T)

names <- labels(data)

meta <- data.frame(cbind(names,loc))

data <- 
  read.dna("TM_Sample.fasta", format = "fasta")

hap <- haplotype(data)

hapInfo <- stack(setNames(attr(hap,"index"),rownames(hap)))
head(hapInfo)

names(hapInfo) <- c("index","haplotype")

merged <- data.frame(cbind(hapInfo,meta[hapInfo$index,]))

net <- haploNet(hap)
plot(net,size=attr(net,"freq"))

pie <- table(merged$haplotype,merged$loc)
head(pie)


svg("COI_TCS_haplotype.svg", width = 7, height = 6)
plot(net,size=attr(net,"freq"),pie=pie)
legend("bottomleft", colnames(pie), col=rainbow(ncol(pie)), pch=19, ncol=2)
dev.off()
```

```{r}
write.nexus.data(data,
                 file = "TM.nex")
```
