---
title: "Photosymbiont genotype and species"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load(tidyverse,
               janitor,
               rstatix,
               ggbeeswarm,
               cld,
               ape,
               ggtree,
               viridis)
```

# Genus

Load data.

```{r}
# ITS2
A.1_ITS2 <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

A.1_ITS2_1 <- 
  A.1_ITS2 |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

# Metadata
A.1_Metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx")

A.1_Metadata_1 <- 
  A.1_Metadata |> 
  filter(SymportalID %in% (A.1_ITS2_1 |> pull(sample_name))) |> 
  rename(sample_name = SymportalID) |> 
  select(1:3, region)

# Merged
A.1_ITS2_2 <- 
  A.1_ITS2_1 |> 
  left_join(A.1_Metadata_1) |> 
  relocate(sample_name, CorrectID, Species, region)

A.1_ITS2_3 <- 
  A.1_ITS2_2 |> 
  pivot_longer(!c(1:4),
               names_to = "ITS2",
               values_to = "Count") |> 
  mutate(Genus = case_when(str_detect(ITS2, "A") ~ "Symbiodinium",
                           str_detect(ITS2, "C") ~ "Cladocopium",
                           str_detect(ITS2, "D") ~ "Durusdinium",
                           .default = "Gerakladium")) |> 
  group_by(Species, CorrectID, Genus) |> 
  summarise(Total = sum(Count)) |> 
  ungroup() |> 
  group_by(Species, CorrectID) |> 
  mutate(RelativeAbundance = Total/sum(Total) * 100) |> 
  ungroup() |> 
  filter(Genus != "Gerakladium") 

# Plot
A.2_ITS2 <- 
  A.1_ITS2_3 |> 
  ggplot(aes(x = Genus,
             y = RelativeAbundance,
             color = Genus,
             fill = Genus)) +
  stat_summary(fun = "mean",
               geom = "bar",
               alpha = 0.75,
               color = NA,
               width = 0.75) +
  stat_summary(fun = mean,
               fun.data = mean_se,
               geom = "linerange")  +
  geom_jitter(width = 0.1, alpha = 0.25) +
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,100)) +
  facet_wrap(~Species,
             axes = "all_y") +
  theme_minimal() +
  coord_cartesian(clip = "off") +
  theme(panel.grid = element_blank(),
        axis.line.y = element_line(),
        axis.ticks.y = element_line(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        legend.text = element_text(face = "italic"),
        legend.position = "right",
        strip.text = element_text(face = "italic")) +
  scale_fill_manual(values = 
                      c("Symbiodinium"= "#746FB4",
                        "Cladocopium" = "#209E77",
                        "Durusdinium" = "#E7AB06",
                        "Gerakladium" = "#A9A9A9")) +
  scale_color_manual(values = 
                      c("Symbiodinium"= "#746FB4",
                        "Cladocopium" = "#209E77",
                        "Durusdinium" = "#E7AB06",
                        "Gerakladium" = "#A9A9A9"))

A.2_ITS2 |> # save
  ggsave(filename = "Bar - Relative abundance among photosymbs within species.svg", 
         dpi = 900,
         width = 17, 
         height = 10,
         unit = "cm")
```

Stats.

```{r}
A.3_ITS2 <- 
  A.1_ITS2_3 |> 
  group_by(Species) |> 
  friedman_test(RelativeAbundance ~ Genus | CorrectID)

A.3_ITS2Pair <- 
  A.1_ITS2_3 |> 
  group_by(Species) |> 
  pairwise_wilcox_test(
    RelativeAbundance ~ Genus, 
    paired = TRUE,
    p.adjust.method = "BH")

A.3_ITS2Pair_1 <- 
  A.1_ITS2_3 %>%
  group_by(Species) %>%
  do({
    # run pairwise Wilcoxon for that species
    res <- pairwise_wilcox_test(
      data = .,
      RelativeAbundance ~ Genus,
      paired = TRUE,
      p.adjust.method = "holm"
    )
    # convert to htest format (or use the data.frame format make_cld accepts)
    # here easiest: use res which is a data.frame with group1, group2, p.adj
    # pass that to make_cld()
    cld_out <- make_cld(res, 
                        p.col = "p.adj", 
                        group1 = "group1", 
                        group2 = "group2",
                        # optionally: threshold = 0.05
                       )
    # add species column
    cld_out %>% mutate(Species = unique(.$Species))
  }) %>%
  ungroup()
```

# ITS2 type profile

```{r}
B.1_ITS2TypeProf <- 
  read_tsv("../raw_data/20250829T091611/its2_type_profiles/4_first_analysis_20250829T091611.profiles.absolute.abund_and_meta.txt")

B.1_ITS2TypeProf_1 <- # tidy
  B.1_ITS2TypeProf |> 
  slice(-(1:5)) %>% # Remove first 5 rows (unnecessary rows)
  row_to_names(1) |>  # Move first row to column header
  rename(sample_name = 2) |>  # Name 2nd column
  select(-1) |>  # Remove first column (unneeded)
  drop_na() |> 
  type_convert() 

B.1_ITS2TypeProf_2 <- # with metadata
  A.1_Metadata_1 |> 
  left_join(B.1_ITS2TypeProf_1) 

B.1_ITS2TypeProf_3 <- # edit manually
  B.1_ITS2TypeProf_2 |> 
  edit()

B.1_ITS2TypeProf_4 <- 
  B.1_ITS2TypeProf_3 |> 
  pivot_longer(!c(1:4),
               names_to = "ITS2TypeProf",
               values_to = "Count") |> 
  group_by(CorrectID) |> 
  mutate(RelAbun = Count/sum(Count)) |> 
  ungroup() |> 
  mutate(Genus = str_extract(ITS2TypeProf, "[A-Z]"))
```

Plot.

```{r}
B.2_ITS2TypeProf <- 
  B.1_ITS2TypeProf_4 |> 
  ggplot(aes(x = CorrectID,
             y = ITS2TypeProf,
             fill = RelAbun)) +
  geom_raster() +
  facet_grid(rows = vars(Genus), 
             cols = vars(Species),
             scale = "free",
             space = "free") +
  scale_fill_viridis(option = "magma")
```

Incorporate distance between ITS2 type profiles.

```{r}
## Cladocopium
# read distance matrix
B.1_Claodcopium <- 
  read_tsv("../raw_data/20250829T091611/between_profile_distances/C/20250829T091611_unifrac_profile_distances_C_sqrt.dist",
           col_names = F) |> 
  select(!2) |> 
  edit()

# convert to matrix
B.1_Claodcopium_1 <- as.matrix(B.1_Claodcopium[,-1])
rownames(B.1_Claodcopium_1) <- B.1_Claodcopium[[1]]

# distance object
B.1_Claodcopium_2 <- as.dist(B.1_Claodcopium_1)

# NJ tree
B.1_Claodcopium_3 <- 
  hclust(B.1_Claodcopium_2,
         method = "average") |> 
  as.phylo()

# ggtree plot
B.1_Claodcopium_4 <- 
  ggtree(B.1_Claodcopium_3) +
  geom_tiplab(size = 3)

B.1_Claodcopium_5 <- # tip order
  B.1_Claodcopium_3$tip.label


## Durusdinium
# read distance matrix
B.1_Durusdinium <- 
  read_tsv("../raw_data/20250829T091611/between_profile_distances/D/20250829T091611_unifrac_profile_distances_D_sqrt.dist",
           col_names = F) |> 
  select(!2) |> 
  edit()

# convert to matrix
B.1_Durusdinium_1 <- as.matrix(B.1_Durusdinium[,-1])
rownames(B.1_Durusdinium_1) <- B.1_Durusdinium[[1]]

# distance object
B.1_Durusdinium_2 <- as.dist(B.1_Durusdinium_1)

# NJ tree
B.1_Durusdinium_3 <- 
  hclust(B.1_Durusdinium_2,
         method = "average") |> 
  as.phylo()

# ggtree plot
B.1_Durusdinium_4 <- 
  ggtree(B.1_Durusdinium_3) +
  geom_tiplab(size = 3)

B.1_Durusdinium_5 <- # tip order
  B.1_Durusdinium_3$tip.label


## Symbiodinium
# read distance matrix
B.1_Symbiodinium <- 
  read_tsv("../raw_data/20250829T091611/between_profile_distances/A/20250829T091611_unifrac_profile_distances_A_sqrt.dist",
           col_names = F) |> 
  select(!2) |> 
  edit()

# convert to matrix
B.1_Symbiodinium_1 <- as.matrix(B.1_Symbiodinium[,-1])
rownames(B.1_Symbiodinium_1) <- B.1_Symbiodinium[[1]]

# distance object
B.1_Symbiodinium_2 <- as.dist(B.1_Symbiodinium_1)

# UPGMA tree
B.1_Symbiodinium_3 <- 
  hclust(B.1_Symbiodinium_2,
         method = "average") |> 
  as.phylo()

# ggtree plot
B.1_Symbiodinium_4 <- 
  ggtree(B.1_Symbiodinium_3) +
  geom_tiplab(size = 3)

B.1_Symbiodinium_5 <- # tip order
  B.1_Symbiodinium_3$tip.label
```

Replot.

```{r}
B.1_MergedArranged <- 
  c(B.1_Symbiodinium_5, B.1_Claodcopium_5, B.1_Durusdinium_5)

B.1_ITS2TypeProf_5 <- 
  B.1_ITS2TypeProf_4 |> 
  mutate(ITS2TypeProf = factor(ITS2TypeProf, levels = B.1_MergedArranged)) %>%
  arrange(ITS2TypeProf) |> 
  mutate(SpeciesXRegion = paste0(Species, "_", region))

B.2_ITS2TypeProf_1 <- 
  B.1_ITS2TypeProf_5 |> 
  ggplot(aes(x = CorrectID,
             y = ITS2TypeProf,
             fill = RelAbun)) +
  geom_raster() +
  facet_grid(rows = vars(Genus), 
             cols = vars(Species),
             scale = "free",
             space = "free") +
  scale_fill_viridis(option = "magma")
```

## Average within species

```{r}
C.1_ITS2TypeProf <- 
  B.1_ITS2TypeProf_4 |> 
  group_by(Species, ITS2TypeProf, Genus) |> 
  summarize(Average = mean(RelAbun)) |> 
  ungroup() |> 
  mutate(ITS2TypeProf = factor(ITS2TypeProf, levels = B.1_MergedArranged)) %>%
  arrange(ITS2TypeProf)

C.2_ITS2TypeProf <- 
  C.1_ITS2TypeProf |> 
  ggplot(aes(x = Species,
             y = ITS2TypeProf,
             fill = Average)) +
  facet_grid(rows = vars(Genus),
             space = "free",
             scales = "free") +
  geom_raster() +
  scale_fill_viridis(option = "magma")
```

## Presence

```{r}
D.1_ITS2TypeProf <- 
  B.1_ITS2TypeProf_4 |> 
  mutate(Presence = if_else(Count > 0, 1, 0)) |> 
  group_by(Species, ITS2TypeProf) |> 
  summarise(TotalPresence = sum(Presence)) |> 
  ungroup() |> 
  mutate(TotalPresence = if_else(TotalPresence > 0, 1, 0)) |> 
  mutate(ITS2TypeProf = factor(ITS2TypeProf, levels = B.1_MergedArranged)) %>%
  arrange(ITS2TypeProf)

D.2_ITS2TypeProf <- 
  D.1_ITS2TypeProf |> 
  ggplot(aes(x = fct_relevel(Species,
                             c("H. hippopus",
                               "H. porcellanus",
                               "T. crocea",
                               "T. squamosa",
                               "T. noae",
                               "T. maxima",
                               "T. derasa", 
                               "T. gigas")),
             y = ITS2TypeProf,
             fill = TotalPresence)) +
#  facet_grid(rows = vars(Genus),
#             cols = vars(Species),
#             space = "free",
#             scales = "free") +
  labs(x = "Species") +
  geom_tile(color = "white") +
  scale_fill_viridis(option = "cividis") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90,
    vjust = 0.5,
    hjust = 1))
```
