---
title: "Photosymbiont composition among regions"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Comm. structure

Load pkgs.

```{r}
pacman::p_load(tidyverse,
               janitor,
               vegan,
               cowplot,
               pairwiseAdonis,
               svglite,
               viridis,
               multcompView,
               ANCOMBC,
               phyloseq,
               scico,
               ggbeeswarm,
               ggupset,
               rstatix)
```

Load metadata.

```{r}
A.1_Metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx")
```

Load ITS.

```{r}
A.1_ITS2 <- read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")
```

Tidy.

```{r}
A.1_Metadata_1 <- 
  A.1_Metadata |> 
  select(c(1, 2, 3, 12)) |> 
  rename(sample_name = 1)

A.1_ITS2_1 <- 
  A.1_ITS2 |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

A.1_ITS2_2 <- 
  A.1_ITS2_1 |> 
  left_join(A.1_Metadata_1) |> 
  relocate(sample_name, CorrectID, Species, region)
```

Make matrix.

```{r}
A.1_ITS2_3 <- # retain CorrectID only
  A.1_ITS2_2 |> 
  select(!c(1,3,4)) |> 
  arrange(CorrectID) |> 
  column_to_rownames("CorrectID") |> 
  as.matrix() |> 
  sqrt()
```

Compute Bray-Curtis dissimilarity.

```{r}
A.1_ITS2BrayCurtis <- 
  A.1_ITS2_3 |> 
  vegdist()
```

Do NMDS.

```{r}
A.1_ITS2BrayCurtisNMDS <- 
  A.1_ITS2BrayCurtis |> 
  metaMDS(k = 2)

A.1_ITS2BrayCurtisNMDS_1 <- # tidy NMDS data
  as.data.frame(scores(A.1_ITS2BrayCurtisNMDS)) |> 
  rownames_to_column("CorrectID")

A.1_ITS2BrayCurtisNMDS_2 <- # add metadata
  A.1_ITS2BrayCurtisNMDS_1 |> 
  left_join(A.1_Metadata_1)
```

Plot.

```{r}
A.2_ITS2BrayCurtisNMDS <- 
  A.1_ITS2BrayCurtisNMDS_2 |> 
  ggplot(aes(x = NMDS1, y = NMDS2)) +
  geom_hline(yintercept = 0,
             color = 'darkgray',
             linetype = 'dashed',
             linewidth = 0.25) +
  geom_vline(xintercept = 0,
             color = 'darkgray',
             linetype = 'dashed',
             linewidth = 0.25) +
  geom_point(aes(color = fct_relevel(region,
                                    c("Alaminos",
                                      "Antique",
                                      "ARNP",
                                      "Batangas",
                                      "Bohol",
                                      "Bolinao",
                                      "Calaguas",
                                      "Camiguin",
                                      "Palawan",
                                      "Samar",
                                      "Zambales"))),
             size = 2) +
  labs(x = "NMDS1",
       y = "NMDS2",
       color = "Region") +
  scale_color_manual(values = 
                       c("Alaminos"="#E41A1C",
                         "Antique"="#377EB8",
                         "ARNP"="#4DAF4A",
                         "Batangas"="#984EA3",
                         "Bohol"="#FF7F00",
                         "Bolinao"="#FFFF33",
                         "Calaguas"="#A65628",
                         "Camiguin"="#F781BF",
                         "Palawan"="#999999",
                         "Samar"="#66C2A5",
                         "Zambales"="#FC8D62")) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.line = element_line(),panel.background    = element_rect(fill = "transparent", colour = NA),
        plot.background     = element_rect(fill = "transparent", colour = NA),
        legend.background   = element_rect(fill = "transparent", colour = NA),
        legend.box.background = element_rect(fill = "transparent", colour = NA),
        legend.position = "bottom")

# Save
A.2_ITS2BrayCurtisNMDS |> 
  ggsave(filename = "NMDS - All regions.svg", 
         dpi = 900,
         width = 17.5, 
         units = "cm")

(A.2_ITS2BrayCurtisNMDS +
  theme(legend.position = "none")) |> 
  ggsave(filename = "NMDS - All regions - No legend.svg", 
         dpi = 900,
         width = 8.5,
         height = 8.5, 
         units = "cm")
```

Plot, separated by regions.

```{r}
A.2_ITS2BrayCurtisNMDS_1 <- 
  A.1_ITS2BrayCurtisNMDS_2 |> 
  ggplot(aes(x = NMDS1, y = NMDS2)) +
  geom_hline(yintercept = 0,
             color = 'darkgray',
             linetype = 'dashed',
             linewidth = 0.25) +
  geom_vline(xintercept = 0,
             color = 'darkgray',
             linetype = 'dashed',
             linewidth = 0.25) +
  stat_ellipse(level = 0.5,
               aes(fill = region),
               geom = "polygon",
               alpha = 0.5) +
  stat_ellipse(level = 0.75,
               aes(fill = region),
               geom = "polygon",
               alpha = 0.25) +
  stat_ellipse(level = 0.9,
               aes(fill = region),
               geom = "polygon",
               alpha = 0.1) +
  geom_point(aes(color = fct_relevel(region,
                                    c("Alaminos",
                                      "Antique",
                                      "ARNP",
                                      "Batangas",
                                      "Bohol",
                                      "Bolinao",
                                      "Calaguas",
                                      "Camiguin",
                                      "Palawan",
                                      "Samar",
                                      "Zambales"))),
             size = 1) +
  labs(x = "NMDS1",
       y = "NMDS2",
       color = "GC species") +
  scale_color_manual(values = 
                       c("Alaminos"="#E41A1C",
                         "Antique"="#377EB8",
                         "ARNP"="#4DAF4A",
                         "Batangas"="#984EA3",
                         "Bohol"="#FF7F00",
                         "Bolinao"="#FFFF33",
                         "Calaguas"="#A65628",
                         "Camiguin"="#F781BF",
                         "Palawan"="#999999",
                         "Samar"="#66C2A5",
                         "Zambales"="#FC8D62")) +
  scale_fill_manual(values = 
                       c("Alaminos"="#E41A1C",
                         "Antique"="#377EB8",
                         "ARNP"="#4DAF4A",
                         "Batangas"="#984EA3",
                         "Bohol"="#FF7F00",
                         "Bolinao"="#FFFF33",
                         "Calaguas"="#A65628",
                         "Camiguin"="#F781BF",
                         "Palawan"="#999999",
                         "Samar"="#66C2A5",
                         "Zambales"="#FC8D62")) +
  facet_wrap(~fct_relevel(region,
                         c("Alaminos",
                           "Antique",
                           "ARNP",
                           "Batangas",
                           "Bohol",
                           "Bolinao",
                           "Calaguas",
                           "Camiguin",
                           "Palawan",
                           "Samar",
                           "Zambales"))) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.line = element_line(),panel.background    = element_rect(fill = "transparent", colour = NA),
        plot.background     = element_rect(fill = "transparent", colour = NA),
        legend.background   = element_rect(fill = "transparent", colour = NA),
        legend.box.background = element_rect(fill = "transparent", colour = NA),
        legend.text = element_text())

# Save
(A.2_ITS2BrayCurtisNMDS_1 +
  theme(legend.position = "none")) |> 
  ggsave(filename = "NMDS - Per region - No legend.svg", 
         dpi = 900,
         width = 8.5,
         height = 8.5, 
         units = "cm")
```

## PERMANOVA

Stats.

```{r}
A.2_PERMANOVA <- 
  adonis2(A.1_ITS2BrayCurtis ~ get("region"), 
          data = (A.1_Metadata_1 |> 
                    filter(CorrectID %in% (A.1_ITS2_2 |>
                                             pull(CorrectID))) |>
                    arrange(CorrectID)))

A.2_PERMANOVAPwise <- # Pairwise PERMANOVA
  pairwise.adonis2(A.1_ITS2BrayCurtis ~ region,
                   data = (A.1_Metadata_1 |>
                             filter(CorrectID %in% (A.1_ITS2_2 |>
                                                      pull(CorrectID))) |>
                      arrange(CorrectID)),
                   nperm = 9999)
```

Script to tidy pairwisePERMANOVA.

```{r}
Z.fun_PPERMANOVACleaner <- 
  function(pmanova_res){
    a <- 
      pmanova_res[-1] |> 
      bind_rows(.id = 'groups') |> 
      drop_na() %>%
      .$`Pr(>F)` |> 
      p.adjust(method = 'BH') 
    
    b <- 
      unique(pmanova_res[-1] |> 
               bind_rows(.id = 'groups'))$groups |> 
      unique()
      
    d <- 
       a %>% 
        data.frame(b,.) |> 
        rename('groups' = 1,
               `P-adjusted` = 2) %>% 
        left_join(pmanova_res[-1] |> 
                    bind_rows(.id = 'groups'), 
                  ., 
                  by = 'groups') |> 
        mutate(`P-adjusted` = case_when(is.na(`Pr(>F)`) ~ NA,
                                        .default = `P-adjusted`)) |> 
        mutate(Parameter = rep(c('Species', 'Residual', 'Total'), length.out = n())) |> 
        relocate(last_col(), .after = 1)
    
    return(d)
  }
```

Tidy.

```{r}
A.2_PERMANOVAPwise_1 <- # tidy
  A.2_PERMANOVAPwise |> 
  Z.fun_PPERMANOVACleaner()

# Save
A.2_PERMANOVAPwise_1 |> 
  writexl::write_xlsx(path = "ST - PERMANOVA - Region.xlsx")

A.2_PERMANOVAPwise_2 <- # tidier (for cld)
  A.2_PERMANOVAPwise_1 |> 
  drop_na() |> 
  mutate(groups = str_replace(groups, "_vs_", "-")) |> 
  select(groups, `P-adjusted`) |> 
  deframe()

A.2_PERMANOVAPwise_3 <- 
  multcompLetters(A.2_PERMANOVAPwise_2, threshold = 0.05)
```

## PERMDISP

```{r}
A.1_PERMDISP <- 
  betadisper(A.1_ITS2BrayCurtis, 
             group = (A.1_Metadata_1 |>
                        filter(CorrectID %in% (A.1_ITS2_2 |>
                                                 pull(CorrectID))))[["region"]]) |> 
  anova()


A.1_PERMDISPPwise <- 
  TukeyHSD(betadisper(A.1_ITS2BrayCurtis, 
             group = (A.1_Metadata_1 |>
                        filter(CorrectID %in% (A.1_ITS2_2 |>
                                                 pull(CorrectID))))[["region"]]))$group |> 
  as.data.frame() |> 
  rownames_to_column('Comparison') |> 
  mutate(diff = abs(diff)) |> 
  separate(Comparison,
           into = c("Group2", "Group1"),
           sep = "-") %>%
  mutate(sig = case_when(
    `p adj` <= 0.05  ~ "bold",
    TRUE ~ "plain"
  )) %>%
  mutate(diff_label = round(diff,2))

# Save
A.1_PERMDISPPwise |> 
  writexl::write_xlsx(path = "ST - PERMDISP - Region.xlsx")

A.1_PERMDISPPwise_1 <- 
  A.1_PERMDISPPwise |> 
  mutate(Comparison = paste0(Group1, "-", Group2)) |> 
  select(Comparison, `p adj`) |> 
  deframe() |> 
  multcompLetters(threshold = 0.05)
```

# Genus

Tidy data.

```{r}
B.1_ITS2 <- # clean its2 data with correct id info
  A.1_ITS2_2

B.1_ITS2_1 <- # pivot longer
  B.1_ITS2 |> 
  select(!c(1,3,4)) |> 
  pivot_longer(!1,
               names_to = "ITS2",
               values_to = "Count")

B.1_ITS2_2 <- # merge with metadata
  B.1_ITS2_1 |> 
  left_join(A.1_Metadata_1)

B.1_ITS2_3 <- # genus info
  B.1_ITS2_2 |> 
  # Detect genus
  mutate(Genus = case_when(str_detect(ITS2, "A") ~ "Symbiodinium",
                           str_detect(ITS2, "C") ~ "Cladocopium",
                           str_detect(ITS2, "D") ~ "Durusdinium",
                           str_detect(ITS2, "G") ~ "Gerakladium",
                           .default = NA)) |> 
  drop_na() |> 
  group_by(CorrectID, Genus, Species, region) |> 
  summarise(Total = sum(Count))
```

Plot.

```{r}
B.2_ITS2Genus <- 
  B.1_ITS2_3 |> 
  ungroup() |> 
  ggplot(aes(x = CorrectID,
             y = Total,
             fill = Genus)) +
  geom_bar(stat = "identity",
           position = "fill",
           width = 1) +
  facet_grid(cols = vars(region),
             space = "free",
             scale = "free") +
  scale_fill_manual(values = 
                      c("Symbiodinium"= "#746FB4",
                        "Cladocopium" = "#209E77",
                        "Durusdinium" = "#E7AB06",
                        "Gerakladium" = "#A9A9A9")) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        strip.background = element_blank(),
        panel.spacing = unit(0.1,'lines'),
        axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank())
```

Cluster per region.

```{r}
B.1_ITS2_4 <- # samples based on ITS2 withing species
  A.1_ITS2_2 |> 
  select(!c(1,3)) |> 
  
  # Prepare for transformation to matrix
  arrange(CorrectID) |> 
  group_by(region) |> 
  group_split(.keep = TRUE) |> 
  map_dfr(~ {
    df <- .x

    # Safely convert to matrix: remove SampleID and any non-numeric columns
    mat_sub <- df %>%
      column_to_rownames("CorrectID") %>%
      select(where(is.numeric)) %>%
      as.matrix() |> 
      sqrt()
    
    dist_mat <- vegdist(mat_sub)  # you can use vegdist() if ecological
    hc <- hclust(dist_mat, method = "ward.D2")
    
    ordered_sample_ids <- .x$CorrectID[hc$order]
    
    tibble(CorrectID = ordered_sample_ids,
           order_within_region = seq_along(ordered_sample_ids),
           region = unique(.x$region))
  }) %>%
  bind_rows()
```

Replot.

```{r}
B.1_ITS2_5 <- 
  B.1_ITS2_3 %>%
  ungroup() |> 
  left_join(B.1_ITS2_4, by = c("CorrectID", "region")) |> 
  arrange(region, order_within_region) |> 
  mutate(
    CorrectID = as.character(CorrectID),
    CorrectID = factor(CorrectID, levels = unique(CorrectID))
  )


B.2_ITS2Genus_1 <- 
  B.1_ITS2_5 |> 
  ggplot(aes(x = CorrectID,
             y = Total,
             fill = fct_relevel(Genus,
                                c("Symbiodinium",
                                  "Cladocopium",
                                  "Durusdinium")))) +
  geom_col(position = "fill",
           width = 1.15,
           color = NA,
           linewidth = 0) +
  facet_grid(cols = vars(fct_relevel(region,
                                     c("Alaminos",
                                       "Antique",
                                       "ARNP",
                                       "Batangas",
                                       "Bohol",
                                       "Bolinao",
                                       "Calaguas",
                                       "Camiguin",
                                       "Palawan",
                                       "Samar",
                                       "Zambales"))),
             space = "free",
             scales = "free_x") +
  scale_fill_manual(values = c(
    "Symbiodinium"= "#746FB4",
    "Cladocopium" = "#209E77",
    "Durusdinium" = "#E7AB06",
    "Gerakladium" = "#A9A9A9"
  )) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    strip.background = element_blank(),
    panel.spacing = unit(0.1,'lines'),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    axis.line.y = element_line(),
    axis.ticks.y = element_line(),
    axis.text = element_blank(),
  )
```

Save.

```{r}
(B.2_ITS2Genus_1 +
    theme(legend.position = "none")) |> 
  ggsave(filename = "Barplot - Genus - Region - No legend.svg", 
         dpi = 900,
         width = 14.75, 
         height = 2,
         unit = "cm")

B.2_ITS2Genus_1 |> 
  ggsave(filename = "Barplot - Genus - Region.svg", 
         dpi = 900,
         width = 20, 
         height = 10,
         unit = "cm")
```

# ITS2

Tidy data.

```{r}
C.1_ITS2 <- # clean its2 data with correct id info
  A.1_ITS2_2

C.1_ITS2_1 <- # pivot longer
  C.1_ITS2 |> 
  select(!c(1,3,4)) |> 
  pivot_longer(!1,
               names_to = "ITS2",
               values_to = "Count")

C.1_ITS2_2 <- # merge with metadata
  C.1_ITS2_1 |> 
  left_join(A.1_Metadata_1)

C.1_ITS2_3 <- # genus info
  C.1_ITS2_2 |> 
  # Detect genus
  mutate(Genus = case_when(str_detect(ITS2, "A") ~ "Symbiodinium",
                           str_detect(ITS2, "C") ~ "Cladocopium",
                           str_detect(ITS2, "D") ~ "Durusdinium",
                           str_detect(ITS2, "G") ~ "Gerakladium",
                           .default = NA)) |> 
  drop_na()
```

Get most abundant seqs.

```{r}
C.1_ITS2Top <- 
  C.1_ITS2_3 |>
  ungroup() |> 
  group_by(ITS2) |>
  summarise(TotalCount = sum(Count)) |> 
  ungroup() |> 
  mutate(RelativeAbundance = round(TotalCount/sum(TotalCount),4)) |> 
  arrange(desc(RelativeAbundance)) |> 
  filter(RelativeAbundance > 0.01) |> # get its2 seqs with >1% relative abundance
  pull(ITS2)
```

Square root-transform.

```{r}
C.1_ITS2_4 <- 
  C.1_ITS2_3 |> 
  mutate(Count = sqrt(Count)) |> 
  filter(ITS2 %in% C.1_ITS2Top) |> 
  ungroup() |> 
  left_join(B.1_ITS2_4, by = c("CorrectID", "region"))  |> 
  arrange(region, order_within_region) |> 
  mutate(
    CorrectID = as.character(CorrectID),
    CorrectID = factor(CorrectID, levels = unique(CorrectID))
  )
```

Plot.

```{r}
C.2_ITS2 <- 
  C.1_ITS2_4 |> 
  ggplot(aes(x = CorrectID, y = ITS2, fill = Count)) +
  geom_raster() +
  scale_fill_viridis(option = "mako", name = "sqrt(count)") +
  facet_grid(vars(region),
             rows = vars(fct_relevel(Genus,
                                     c("Symbiodinium",
                                       "Cladocopium",
                                       "Durusdinium"))),
             scales = "free",
             space = "free") +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        panel.spacing = unit(0.1,'lines'),
        panel.grid = element_blank(),
        axis.text.y = element_text(size = 6, margin = margin(r = -1)),
        strip.text.y = element_blank())
```

Save.

```{r}
(C.2_ITS2 +
    theme(legend.position = "none",
          axis.text = element_blank())) |> 
  ggsave(filename = "Heatmap - ITS2 - Region - No legend.svg", 
         dpi = 900,
         width = 15.7, 
         height = 3.75,
         unit = "cm")

C.2_ITS2 |> 
  ggsave(filename = "Heatmap - ITS2 - Region.svg", 
         dpi = 900,
         width = 17.5, 
         height = 20,
         unit = "cm")
```

## DA

Make ANCOMBC2-readable. Make a phyloseq table.

```{r}
C.1_ITS2ANCOMTable <- 
  C.1_ITS2 |> 
  select(!c(1,3,4)) |> # pivot longer
  pivot_longer(!CorrectID,
              names_to = "ITS2",
              values_to = "Count") |>  # pivot wider
  pivot_wider(names_from = CorrectID,
              values_from = Count) |> 
  column_to_rownames("ITS2") |> 
  as.matrix() |> 
  otu_table(taxa_are_rows = T)
```

Phyloseq taxonomy mapping.

```{r}
C.1_ITS2ANCOMTaxa <-
  C.1_ITS2 |> 
  select(!c(1,3,4)) |> # pivot longer
  pivot_longer(!CorrectID,
              names_to = "ITS2",
              values_to = "Count") |>  # pivot wider
  pivot_wider(names_from = CorrectID,
              values_from = Count) |> 
  mutate(Phylum = ITS2,
         Class = NA,
         Order = NA,
         Family = NA,
         Genus = NA,
         Species = NA) |> 
  select(1, Phylum:Species) |> 
  column_to_rownames('ITS2') |> 
  as.matrix() |> 
  tax_table()
```

Phyloseq metadata.

```{r}
C.1_ITS2ANCOMMeta <- 
  A.1_Metadata_1 |> 
  filter(CorrectID %in% (A.1_ITS2_2 |> 
                           pull(CorrectID))) |> 
  select(!sample_name) |> 
  column_to_rownames("CorrectID") |> 
  as.data.frame() |> 
  sample_data()
```

Phyloseq merge.

```{r}
C.1_ITS2ANCOMMerge <- 
  phyloseq(C.1_ITS2ANCOMTable,
           C.1_ITS2ANCOMTaxa)

C.1_ITS2ANCOMMerge_1 <- 
  merge_phyloseq(C.1_ITS2ANCOMMerge,
                 C.1_ITS2ANCOMMeta)
```

ANCOMBC2 among regions.

```{r}
C.1_ITS2ANCOMRun <- ANCOMBC::ancombc2(
  data = C.1_ITS2ANCOMMerge_1,
  fix_formula = "region",
  rand_formula = NULL,
  group = "region",
  global = TRUE,
  pairwise = TRUE,
  tax_level = "Phylum",
  p_adj_method = "BH",
  lib_cut = 1000,
  prv_cut = 0.10,
  alpha = 0.05,
  verbose = TRUE
)
```

Tidy.

```{r}
C.1_ITS2ANCOMRunRes <- 
  C.1_ITS2ANCOMRun$res

C.1_ITS2ANCOMRunGlobal <- 
  C.1_ITS2ANCOMRun$res_global

C.1_ITS2ANCOMRunPairwise <- 
  C.1_ITS2ANCOMRun$res_pair

C.1_ITS2ANCOMRunPairwise_Z <- 
  C.1_ITS2ANCOMRunPairwise |> 
  pivot_longer(
    cols = -taxon,
    names_to = c("parameter", "group"),
    names_pattern = "(.+?)_(region.+)",
    values_to = "value")  |> 
  pivot_wider(
    names_from = parameter,
    values_from = value) |> 
  separate(group, sep = "_", into = c("Group1", "Group2")) |> 
  mutate(Group2.5 = if_else(is.na(Group2), str_sub(Group1, 7, -1), NA),
         Group1 = if_else(is.na(Group2), "Alaminos", Group1),
         Group2 = if_else(is.na(Group2), Group2.5, Group2)) |> 
  mutate(Group1 = str_remove_all(Group1, "region"),
         Group2 = str_remove_all(Group2, "region")) |> 
  select(-last_col())

# Save
C.1_ITS2ANCOMRunPairwise_Z |> 
  writexl::write_xlsx(path = "ST - ANCOMBC2 - Region.xlsx")
```

Tidy.

```{r}
C.1_ITS2ANCOMRunGlobal_1 <- 
  C.1_ITS2ANCOMRunGlobal |> 
  filter(diff_abn == TRUE & passed_ss == TRUE)

C.1_ITS2ANCOMRunPairwise_1 <- # diff abundant
  C.1_ITS2ANCOMRunPairwise |> 
  filter(taxon %in% (C.1_ITS2ANCOMRunGlobal_1 |> 
                       pull(taxon))) |> 
  select(taxon, starts_with("diff_")) |> 
  pivot_longer(!taxon,
               names_to = "Comparison",
               values_to = "DiffAbun") |> 
  mutate(Comparison = str_remove(Comparison, "diff_"))

C.1_ITS2ANCOMRunPairwise_2 <- # lfc 
  C.1_ITS2ANCOMRunPairwise |> 
  filter(taxon %in% (C.1_ITS2ANCOMRunGlobal_1 |> 
                       pull(taxon))) |> 
  select(taxon, starts_with("lfc_")) |> 
  pivot_longer(!taxon,
               names_to = "Comparison",
               values_to = "LFC") |> 
  mutate(Comparison = str_remove(Comparison, "lfc_"))

C.1_ITS2ANCOMRunPairwise_3 <- # merge
  C.1_ITS2ANCOMRunPairwise_2 |> 
  left_join(C.1_ITS2ANCOMRunPairwise_1)

C.1_ITS2ANCOMRunPairwise_4 <- # last tidy
  C.1_ITS2ANCOMRunPairwise_3 |> 
  mutate(Comparison = if_else(str_detect(Comparison, "_"), 
                              str_remove_all(Comparison, "region"),
                              str_replace_all(Comparison, "region", "Alaminos_"))) |> 
  separate(col = Comparison,       # replace with your column name
           into = c("Species1", "Species2"),
           sep = "_") |> 
  mutate(Species1 = str_remove_all(str_sub(Species1, 1,8), ". "),
         Species2 = str_remove_all(str_sub(Species2, 1,8), ". ")) |> 
  unite(col = "Comparison", Species1, Species2, sep = "-\n")

C.1_ITS2ANCOMRunPairwise_5 <- # filter
  C.1_ITS2ANCOMRunPairwise_4 |> 
  filter(DiffAbun == TRUE)

C.1_ITS2ANCOMRunPairwise_6 <- 
  C.1_ITS2ANCOMRunPairwise_5 |> 
  complete(taxon = unique(taxon), 
           Comparison = unique(Comparison), 
           fill = list(LFC = 0, DiffAbun = FALSE))
```

No differentially abundant sequences.

# ITS2 type profile

Load data.

```{r}
D.1_ITS2TypeProf <- 
  read_tsv("../raw_data/20250829T091611/its2_type_profiles/4_first_analysis_20250829T091611.profiles.absolute.abund_and_meta.txt")

D.1_ITS2TypeProf_1 <- # tidy
  D.1_ITS2TypeProf |> 
  slice(-(1:5)) %>% # Remove first 5 rows (unnecessary rows)
  row_to_names(1) |>  # Move first row to column header
  rename(sample_name = 2) |>  # Name 2nd column
  select(-1) |>  # Remove first column (unneeded)
  drop_na() |> 
  type_convert() 

D.1_ITS2TypeProf_2 <- # merge with metadata, remove unneeded
  D.1_ITS2TypeProf_1 |> 
  left_join(A.1_Metadata_1) |> 
  filter(sample_name %in% (A.1_ITS2_1 |> pull(sample_name)))

D.1_ITS2TypeProf_3 <- # pivot longer
  D.1_ITS2TypeProf_2 |> 
  pivot_longer(!c(sample_name, CorrectID, Species, region),
               names_to = "ITS2TypeProf",
               values_to = "Count")

D.1_ITS2TypeProf_4 <- # convert counts to 0 or 1
  D.1_ITS2TypeProf_3 |> 
  mutate(Presence = if_else(Count > 0, 1, 0))

D.1_ITS2TypeProf_5 <- # number of ITS2TypeProf per sample
  D.1_ITS2TypeProf_4 |> 
  group_by(CorrectID, region) |> 
  summarise(NoITS2TypeProf = sum(Presence))

D.1_ITS2TypeProf_6 <- # arrange based on clusters
  D.1_ITS2TypeProf_5 |> 
  left_join(B.1_ITS2_4, by = c("CorrectID", "region")) |> 
  ungroup() |> 
  arrange(region, order_within_region) |> 
  mutate(
    CorrectID = as.character(CorrectID),
    CorrectID = factor(CorrectID, levels = unique(CorrectID))
  )
```

Plot.

```{r}
D.2_ITS2TypeProf <- 
  D.1_ITS2TypeProf_6 |> 
  ggplot(aes(x = CorrectID, y = NoITS2TypeProf)) +
  geom_col(width = 1.15,
           fill = "black") +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0,6),
                     breaks = seq(0,6, by = 2)) +
  facet_grid(cols = vars(region),
             scales = "free",
             space = "free") +
  theme_minimal() +
  theme(axis.ticks.y = element_line(color = "black"),
        panel.grid = element_blank(),
        axis.line.y = element_line(color = "black"),
        axis.text.x = element_blank(),
        panel.spacing = unit(0.1,'lines'),
        axis.text.y = element_text(size = 6),
        axis.title = element_blank())

# Save
D.2_ITS2TypeProf |> 
  ggsave(filename = "Barplots - Number of ITS2 type profiles - Region.svg", 
         dpi = 900,
         width = 15, 
         height = 2.25,
         unit = "cm")
```

## Diversity

```{r}
D.1_ITS2TypeProfAlpha <- 
  D.1_ITS2TypeProf_4 |> 
  group_by(CorrectID) |> 
  summarise(Shannon = diversity(Count, index = "shannon"),
            Richness = specnumber(Count)) |> 
  ungroup() |> 
  left_join((A.1_Metadata_1 |> 
               select(c(CorrectID, region))))
```

Plot.

```{r}
D.2_ITS2TypeProfAlpha <- 
  D.1_ITS2TypeProfAlpha |> 
  ggplot(aes(x = region,
             y = Richness)) +
  geom_violin(trim = F,
              fill = "lightgray",
              color = NA,
              alpha = 0.75) +
  geom_boxplot(fill = '#555555',
               width = 0.05,
               outlier.shape = NA,
               color = '#555555') +
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,5)) +
  stat_summary(fun = "median", 
               fill = 'white',
               color = 'black',
               size = 2.5,
               geom = 'point',
               shape = 21,
               stroke = 1) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.line.y = element_line(),
        axis.ticks.y = element_line(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

D.2_ITS2TypeProfAlpha_1 <- 
  D.1_ITS2TypeProfAlpha |> 
  ggplot(aes(x = region,
             y = Shannon)) +
  geom_violin(trim = F,
              fill = "lightgray",
              color = NA,
              alpha = 0.75) +
  geom_boxplot(fill = '#555555',
               width = 0.05,
               outlier.shape = NA,
               color = '#555555') +
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,1)) +
  stat_summary(fun = "median", 
               fill = 'white',
               color = 'black',
               size = 2.5,
               geom = 'point',
               shape = 21,
               stroke = 1) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.line.y = element_line(),
        axis.ticks.y = element_line(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

D.2_ITS2TypeProfAlpha_2 <- 
  plot_grid(D.2_ITS2TypeProfAlpha,
            D.2_ITS2TypeProfAlpha_1,
            ncol = 2)
```

Save.

```{r}
D.2_ITS2TypeProfAlpha_2 |> 
  ggsave(filename = "Alpha.svg", 
         dpi = 900,
         width = 17, 
         height = 5,
         unit = "cm")
```

## Upset

Tidy.

```{r}
D.1_ITS2TypeProfUpset <- 
  D.1_ITS2TypeProf_4 |> 
  group_by(ITS2TypeProf,
           region) |> 
  summarise(TotalCount = sum(Count)) |> 
  ungroup() |> 
  pivot_wider(names_from = ITS2TypeProf,
              values_from = TotalCount) |>  
  mutate_if(is.numeric, as.logical) |> 
  pivot_longer(!region,
               names_to = 'ITS2TypeProf',
               values_to = 'Logic') |> 
  filter(Logic) %>% 
  select(-Logic) %>% 
  group_by(ITS2TypeProf)%>% 
  summarize(region = list(region))
```

Plot.

```{r}
D.2_ITS2TypeProfUpset <-
  D.1_ITS2TypeProfUpset |> 
  ggplot(aes(x = region)) +
  geom_bar() +
  scale_x_upset() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1),
                                        add = 0))  +
  geom_text(aes(label=after_stat(count)),
            stat='Count',
            vjust=-1,
            size = 2.5) +
  theme_classic() +
  theme(panel.grid = element_blank(),
        axis.title = element_blank(),
        plot.margin = margin(0.5,0,0,1,
                             unit ='cm'),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank())
```

Save.

```{r}
D.2_ITS2TypeProfUpset |> 
  ggsave(filename = "Upset.svg", 
         dpi = 900,
         width = 17, 
         height = 8,
         unit = "cm")
```

Stats.

```{r}
# Richness
D.2_ITS2TypeProfAlphaKWRichness <- 
  D.1_ITS2TypeProfAlpha |> 
  kruskal_test(Richness ~ region)

# Shannon
D.2_ITS2TypeProfAlphaKWShannon <- 
  D.1_ITS2TypeProfAlpha |> 
  kruskal_test(Shannon ~ region)
```

# Species labels

Tidy.

```{r}
E.1_Metadata <- # filter metadata to include samples with ITS2 seqs
  A.1_Metadata_1 |> 
  filter(CorrectID %in% (A.1_ITS2_2 |> 
                           pull(CorrectID)))

E.1_Metadata_1 <- # add counts which will be used to do barplots/labels
  E.1_Metadata |> 
  mutate(Count = 1)

E.1_Metadata_2 <- # arrange
  E.1_Metadata_1 |> 
  left_join(B.1_ITS2_4, by = c("CorrectID", "region")) |> 
  arrange(region, order_within_region) |> 
  mutate(
    CorrectID = as.character(CorrectID),
    CorrectID = factor(CorrectID, levels = unique(CorrectID))
  )
```

Plot.

```{r}
E.2_Region <- 
  E.1_Metadata_2 |> 
  ggplot(aes(x = CorrectID,
             y = Count,
             fill = Species)) +
  geom_col(width = 1.15) +
  facet_grid(cols = vars(fct_relevel(region,
                                     c("Alaminos",
                                       "Antique",
                                       "ARNP",
                                       "Batangas",
                                       "Bohol",
                                       "Bolinao",
                                       "Calaguas",
                                       "Camiguin",
                                       "Palawan",
                                       "Samar",
                                       "Zambales"))),
             space = "free",
             scales = "free_x") +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  scale_fill_manual(values = c("H. hippopus"="darkred",
                         "H. porcellanus"="#cf202e",
                         "T. crocea"="#f78822",
                         "T. squamosa"="#f6db35",
                         "T. noae"="#6dc067",
                         "T. maxima"="#4185be",
                         "T. derasa"="#8f499c",
                         "T. gigas"="#201146")) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        panel.spacing = unit(0.1,'lines'),
        axis.title = element_blank(),
        legend.text = element_text(face = "italic"),
        legend.position = "top") +
  guides(fill = guide_legend(nrow = 1))
```

Save.

```{r}
(E.2_Region +
    theme(legend.position = "none",
          axis.text = element_blank()))  |> 
  ggsave(filename = "Labels - Species - No legend.svg", 
         dpi = 900,
         width = 17.5, 
         height = 4,
         unit = "cm")

E.2_Region |> 
  ggsave(filename = "Labels - Species.svg", 
         dpi = 900,
         width = 25, 
         height = 4,
         unit = "cm")
```
