---
title: "GC COI ML tree"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load(ggtree,
               tidyverse,
               treeio,
               ape)
```

# Load data

## Load prior work

```{r}
load("GC COI ML tree.RData")
```

## Load metadata

```{r}
A.1_meta <- 
  read_csv("../Sequence metadata_202510.csv")
```

## Load tree

```{r}
A.tree <- 
  read.iqtree("GCCOISeqs_AlignedAllAddKeep_20251022_clipkitsmartgap.fa.treefile")
```

# Edit tree

```{r}
B.1_Tree <- 
  A.tree |> 
  as_tibble() |> 
  mutate(label = if_else(str_detect(label, "\\|"), # only if it has "|"
                         str_extract(label, "(?<=\\|)[^|]+$"), # extract after the last "|"
                         label)) |>  # otherwise keep as is
  mutate(SampleID = label) |> 
  left_join(A.1_meta) |> 
  mutate(Species = str_replace(Species, pattern = "Tridacna", "T."),
         Species = str_replace(Species, pattern = "Hippopus", "H.")) |> 
  mutate(label = if_else(!is.na(Species),
                         paste(Species, SampleID, sep = " "),
                         SampleID))
```

# Plot

```{r}
B.1_Tree_1 <- 
  B.1_Tree |> 
  as.treedata()

B.2_Tree <- 
  B.1_Tree_1 |> 
  ggtree(layout = "rectangular") +
  geom_tiplab(size = 2)

B.2_Tree_1 <- 
  B.1_Tree_1 |> 
  ggtree(layout = "rectangular",
         branch.length = "none") +
  geom_nodelab(aes(label = node),
               size = 1,
               nudge_x = 0.4) +
  geom_tiplab(size = 1.5) + 
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(10, 50, 10, 10))

ggsave(plot = B.2_Tree_1,
       filename = "Tree01.svg",
       width = 10,
       height = 50,
       unit = "cm")
```

# Reroot

```{r}
C.1_Tree <- 
  B.1_Tree_1 |> 
  root(node = 360,
       resolve.root = T)

C.1_Tree_1 <- 
  C.1_Tree |>
  as_tibble() |> 
  mutate(label = if_else(!is.na(Species),
                         paste(Species, SampleID, sep = " "),
                         SampleID))

C.1_Tree_2 <- 
  C.1_Tree_1 |> 
  as.treedata()

C.2_Tree <- 
  C.1_Tree_2 |> 
  ggtree(layout = "rectangular",
         branch.length = "none") +
  geom_nodelab(aes(label = node),
               size = 1,
               nudge_x = 0.4) +
  geom_tiplab(size = 1.5) + 
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(10, 50, 10, 10))

C.2_Tree <- 
  C.2_Tree |> 
  ggtree::rotate(216)

C.2_Tree <- 
  C.2_Tree |> 
  ggtree::rotate(429)

C.2_Tree <- 
  C.2_Tree |> 
  ggtree::rotate(360)


C.2_Tree <- 
  C.2_Tree |> 
  ggtree::rotate(253)

ggsave(plot = C.2_Tree,
       filename = "Tree02.svg",
       width = 10,
       height = 50,
       unit = "cm")
```

# Finalize

```{r}
D.2_Tree <- 
  C.1_Tree_2 |> 
  ggtree(layout = "rectangular",
         branch.length = "none") +
  geom_nodelab(size = 1,
               nudge_x = 1) +
  geom_tiplab(size = 1.5) + 
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(10, 50, 10, 10))

D.2_Tree <- 
  D.2_Tree |> 
  ggtree::rotate(216)

D.2_Tree <- 
  D.2_Tree |> 
  ggtree::rotate(429)

D.2_Tree <- 
  D.2_Tree |> 
  ggtree::rotate(360)


D.2_Tree <- 
  D.2_Tree |> 
  ggtree::rotate(253)

ggsave(plot = D.2_Tree,
       filename = "Tree03.svg",
       width = 10,
       height = 50,
       unit = "cm")
```

# Collapsed

```{r}
E.2_TreeCollapsed <- 
  D.2_Tree |> 
  ggtree::collapse(361, mode = "mixed", fill = "darkred", color = "darkred") |> 
  ggtree::collapse(383, mode = "mixed", fill = "#cf202e", color = "#cf202e") |> 
  ggtree::collapse(290, mode = "mixed", fill = "#f78822", color = "#f78822") |> 
  ggtree::collapse(254, mode = "mixed", fill = "#f6db35", color = "#f6db35") |> 
  ggtree::collapse(319, mode = "mixed", fill = "#6dc067", color = "#6dc067") |> 
  ggtree::collapse(248, mode = "mixed", fill = "#4185be", color = "#4185be") |> 
  ggtree::collapse(339, mode = "mixed", fill = "#8f499c", color = "#8f499c") |> 
  ggtree::collapse(387, mode = "mixed", fill = "#201146", color = "#201146")

E.vec_NodeToRemove <- 
  E.2_TreeCollapsed |> 
  offspring(419) |> 
  filter(is.na(SH_aLRT)) |> 
  pull(label)

E.vec_TipsToRemove <- 
  c(E.vec_NodeToRemove, "T. mbalavuana KU058677")

E.2_TreeCollapsed_1 <- 
  E.2_TreeCollapsed %>%
  .$data |> 
  as.phylo() |> 
  drop.tip(E.vec_TipsToRemove)

E.2_TreeCollapsed_2 <- 
  E.2_TreeCollapsed_1 |> 
  ggtree(layout = "rectangular") + 
  coord_cartesian(clip = "off")

E.2_TreeCollapsed_3 <- 
  E.2_TreeCollapsed_2 |> 
  ggtree::rotate(205) |> 
  ggtree::rotate(206) |> 
  ggtree::rotate(245) |> 
  ggtree::rotate(381) |> 
  ggtree::rotate(246) |> 
  ggtree::rotate(282)

E.2_TreeCollapsed_4 <- 
  E.2_TreeCollapsed_3 |> 
  ggtree::collapse(382, mode = "mixed", fill = "darkred", color = "darkred") |> 
  ggtree::collapse(404, mode = "mixed", fill = "#cf202e", color = "#cf202e") |> 
  ggtree::collapse(282, mode = "mixed", fill = "#f78822", color = "#f78822") |> 
  ggtree::collapse(246, mode = "mixed", fill = "#f6db35", color = "#f6db35") |> 
  ggtree::collapse(311, mode = "mixed", fill = "#6dc067", color = "#6dc067") |> 
  ggtree::collapse(209, mode = "mixed", fill = "#4185be", color = "#4185be") |> 
  ggtree::collapse(329, mode = "mixed", fill = "#8f499c", color = "#8f499c") |> 
  ggtree::collapse(349, mode = "mixed", fill = "#201146", color = "#201146")

E.2_TreeCollapsed_5 <- 
  E.2_TreeCollapsed_3 + 
  layout_fan(angle = 0) +
  geom_tiplab(size = 1.5, aes(label = SampleID))
```

Save.

```{r}
ggsave(plot = E.2_TreeCollapsed_3,
       filename = "Tree04.svg",
       width = 10,
       height = 30,
       unit = "cm")

ggsave(plot = E.2_TreeCollapsed_4,
       filename = "Tree04 - Collapsed.svg",
       width = 4.5,
       height = 6.5,
       unit = "cm")


ggsave(plot = E.2_TreeCollapsed_5,
       filename = "Tree04 - Uncollapsed.svg",
       width = 17,
       height = 17,
       unit = "cm")
```

# Tree

Initial tree.

```{r}
A.2_tree <- 
  A.tree |> 
  ggtree() +
  geom_tiplab()

reroot(A.2_tree)
```

Reroot.

```{r}
B.tree <- 
  A.tree |> 
  as.phylo() |> 
  root(outgroup = "MW278344.1", resolve.root = TRUE)
```

Rooted tree.

```{r}
B.2_tree <- 
  B.tree |> 
  ggtree(layout = "fan", 
         open.angle = 90, 
         branch.length = "none")
```

Change color.

```{r}
Z.vec_col.spe <-
  c("Tridacna noae"="#5E4FA2",
    "Tridacna squamosa"="#3288BD",
    "Hippopus porcellanus"="#66C2A5",
    "Hippopus hippopus"="#ABDDA4",
    "Tridacna derasa"="#FDAE61",
    "Tridacna gigas"="#F46D43",
    "Tridacna maxima"="#D53E4F",
    "Tridacna crocea"="#9E0142")


C.1_meta <- 
  A.1_meta |> 
  rename(label = SampleID)

C.2_tree <- 
  B.tree |> 
  ggtree(layout = "fan",
         open.angle = 90,
         branch.length = "none")

C.2_tree_1 <- C.2_tree %<+% C.1_meta

C.2_tree_2 <- 
  C.2_tree_1  +
  geom_tippoint(aes(color = Species)) +
  scale_color_manual(values = Z.vec_col.spe)

###
C.2_tree <- 
  B.tree |> 
  ggtree(layout = "rectangular",
         branch.length = "none")

C.2_tree_1 <- C.2_tree %<+% C.1_meta

C.2_tree_3 <- 
  C.2_tree_1 +
  geom_tiplab(size = 2,
              offset = 0.5) +
  geom_tippoint(aes(color = Species),show.legend = F) +
  scale_color_manual(values = Z.vec_col.spe) +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(6, 40, 6, 6, "pt"))

C.2_tree_3 |> 
  ggsave(filename = 'tree_20250725.png',
         dpi = 900,
         width = 2500,
         height = 12000,
         units = "px")

# Edit
C.1_tree <- 
  B.tree |> 
  as_tibble() |> 
  mutate(SampleID = label) |> 
  left_join(A.1_meta, by = "SampleID")

# Plot
C.2_tree <- 
  C.1_tree |> 
  as.treedata() |> 
  ggtree(aes(color = Species),
         layout = "fan",
         open.angle = 90,
         branch.length = "none") +
  scale_color_manual(values = Z.vec_col.spe)
```
