---
title: "Indicator species analysis - Tgig"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load("tidyverse",
               "indicspecies",
               "vegan",
               "compositions",
               "janitor")
```

# ITS2 sequences

Load data.

```{r}
# ITS2 data
A.1_ITS2 <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

A.1_ITS2_1 <- 
  A.1_ITS2 |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

# Metadata
A.1_Metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx")

A.1_Metadata_1 <- # rename col ID
  A.1_Metadata  |> 
  select(c(1, 2, 3, 12)) |> 
  rename(sample_name = 1)

A.1_ITS2_2 <- 
  A.1_ITS2_1 |> 
  left_join(A.1_Metadata_1) |> 
  relocate(sample_name, CorrectID, Species, region)

A.1_Metadata_2 <- # filter metadata
  A.1_Metadata_1 |>
  filter(CorrectID %in% (A.1_ITS2_2 |> pull(CorrectID)))
```

Tidy data. Do not filter.

```{r}
B.1_ITS2 <- 
  A.1_ITS2_2 |> 
  arrange(CorrectID) |> 
  select(-c(1,3,4)) |> 
  column_to_rownames("CorrectID")
```

Filter to get the DIVs only.

```{r}
C.1_DIVs <- 
  read_tsv("../DIVs.txt",col_names = F) |> 
  pull()

C.1_DIVs_1 <- 
  A.1_ITS2_2 |> 
  pivot_longer(-c(1:4),
               names_to = "ITS2",
               values_to = "Count") |> 
  filter(ITS2 %in% C.1_DIVs) |> 
  pivot_wider(names_from = "ITS2",
              values_from = "Count") |> 
  arrange(CorrectID) |> 
  select(-c(1,3,4)) |> 
  column_to_rownames("CorrectID")

# Get Tgigs only
C.1_DIVs_2 <- 
  C.1_DIVs_1 |> 
  rownames_to_column("CorrectID") |> 
  filter(str_detect(CorrectID, pattern = "TG"))

C.1_DIVs_3 <- # pivot longer
  C.1_DIVs_2 |> 
  pivot_longer(-1, names_to = "DIVs", values_to = "Count")

C.1_DIVs_4 <- # get present DIVs
  C.1_DIVs_3 |> 
  group_by(DIVs) |> 
  summarise(Sum = sum(Count)) |> 
  ungroup() |> 
  filter(Sum > 0) |> 
  pull(DIVs)

C.1_DIVs_5 <- # filter
  C.1_DIVs_3 |>
  filter(DIVs %in% C.1_DIVs_4)
```

# Comparison between regions

Bolinao vs. Alaminos

```{r}
D.1_MetadataBolVsAla <- 
  A.1_Metadata_2 |> 
  filter(Species == "T. gigas",
         region %in% c("Bolinao", "Alaminos")) |> 
  arrange(CorrectID)

D.1_DIVBolVsAla <- 
  C.1_DIVs_3 |> 
  filter(CorrectID %in% (D.1_MetadataBolVsAla |> pull(CorrectID))) |> 
  pivot_wider(names_from = "DIVs",
              values_from = "Count") |> 
  arrange(CorrectID) |> 
  column_to_rownames("CorrectID")

D.1_MetadataBolVsAla_1 <- 
  D.1_MetadataBolVsAla |> 
  pull(region) |> 
  factor()

D.1_ISABolVsAla_1 <- 
  multipatt(D.1_DIVBolVsAla, D.1_MetadataBolVsAla_1, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(D.1_ISABolVsAla_1)
```

Bolinao vs. Antique.

```{r}
D.1_MetadataBolVsAnt <- 
  A.1_Metadata_2 |> 
  filter(Species == "T. gigas",
         region %in% c("Bolinao", "Antique")) |> 
  arrange(CorrectID)

D.1_DIVBolVsAnt <- 
  C.1_DIVs_3 |> 
  filter(CorrectID %in% (D.1_MetadataBolVsAnt |> pull(CorrectID))) |> 
  pivot_wider(names_from = "DIVs",
              values_from = "Count") |> 
  arrange(CorrectID) |> 
  column_to_rownames("CorrectID")

D.1_MetadataBolVsAnt_1 <- 
  D.1_MetadataBolVsAnt |> 
  pull(region) |> 
  factor()

D.1_ISABolVsAnt_1 <- 
  multipatt(D.1_DIVBolVsAnt, D.1_MetadataBolVsAnt_1, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(D.1_ISABolVsAnt_1)
```

Bolinao vs. Batangas.

```{r}
D.1_MetadataBolVsBat <- 
  A.1_Metadata_2 |> 
  filter(Species == "T. gigas",
         region %in% c("Bolinao", "Batangas")) |> 
  arrange(CorrectID)

D.1_DIVBolVsBat <- 
  C.1_DIVs_3 |> 
  filter(CorrectID %in% (D.1_MetadataBolVsBat |> pull(CorrectID))) |> 
  pivot_wider(names_from = "DIVs",
              values_from = "Count") |> 
  arrange(CorrectID) |> 
  column_to_rownames("CorrectID")

D.1_MetadataBolVsBat_1 <- 
  D.1_MetadataBolVsBat |> 
  pull(region) |> 
  factor()

D.1_ISABolVsBat_1 <- 
  multipatt(D.1_DIVBolVsBat, D.1_MetadataBolVsBat_1, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(D.1_ISABolVsBat_1)
```

Bolinao vs. Bohol.

```{r}
D.1_MetadataBolVsBoh <- 
  A.1_Metadata_2 |> 
  filter(Species == "T. gigas",
         region %in% c("Bolinao", "Bohol")) |> 
  arrange(CorrectID)

D.1_DIVBolVsBoh <- 
  C.1_DIVs_3 |> 
  filter(CorrectID %in% (D.1_MetadataBolVsBoh |> pull(CorrectID))) |> 
  pivot_wider(names_from = "DIVs",
              values_from = "Count") |> 
  arrange(CorrectID) |> 
  column_to_rownames("CorrectID")

D.1_MetadataBolVsBoh_1 <- 
  D.1_MetadataBolVsBoh |> 
  pull(region) |> 
  factor()

D.1_ISABolVsBoh_1 <- 
  multipatt(D.1_DIVBolVsBoh, D.1_MetadataBolVsBoh_1, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(D.1_ISABolVsBoh_1)
```

Bolinao vs. Camiguin.

```{r}
D.1_MetadataBolVsCam <- 
  A.1_Metadata_2 |> 
  filter(Species == "T. gigas",
         region %in% c("Bolinao", "Camiguin")) |> 
  arrange(CorrectID)

D.1_DIVBolVsCam <- 
  C.1_DIVs_3 |> 
  filter(CorrectID %in% (D.1_MetadataBolVsCam |> pull(CorrectID))) |> 
  pivot_wider(names_from = "DIVs",
              values_from = "Count") |> 
  arrange(CorrectID) |> 
  column_to_rownames("CorrectID")

D.1_MetadataBolVsCam_1 <- 
  D.1_MetadataBolVsCam |> 
  pull(region) |> 
  factor()

D.1_ISABolVsCam_1 <- 
  multipatt(D.1_DIVBolVsCam, D.1_MetadataBolVsCam_1, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(D.1_ISABolVsCam_1)
```

Bolinao vs. Palawan.

```{r}
D.1_MetadataBolVsPal <- 
  A.1_Metadata_2 |> 
  filter(Species == "T. gigas",
         region %in% c("Bolinao", "Palawan")) |> 
  arrange(CorrectID)

D.1_DIVBolVsPal <- 
  C.1_DIVs_3 |> 
  filter(CorrectID %in% (D.1_MetadataBolVsPal |> pull(CorrectID))) |> 
  pivot_wider(names_from = "DIVs",
              values_from = "Count") |> 
  arrange(CorrectID) |> 
  column_to_rownames("CorrectID")

D.1_MetadataBolVsPal_1 <- 
  D.1_MetadataBolVsPal |> 
  pull(region) |> 
  factor()

D.1_ISABolVsPal_1 <- 
  multipatt(D.1_DIVBolVsPal, D.1_MetadataBolVsPal_1, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(D.1_ISABolVsPal_1)
```

Bolinao vs. Samar.

```{r}
D.1_MetadataBolVsSam <- 
  A.1_Metadata_2 |> 
  filter(Species == "T. gigas",
         region %in% c("Bolinao", "Samar")) |> 
  arrange(CorrectID)

D.1_DIVBolVsSam <- 
  C.1_DIVs_3 |> 
  filter(CorrectID %in% (D.1_MetadataBolVsSam |> pull(CorrectID))) |> 
  pivot_wider(names_from = "DIVs",
              values_from = "Count") |> 
  arrange(CorrectID) |> 
  column_to_rownames("CorrectID")

D.1_MetadataBolVsSam_1 <- 
  D.1_MetadataBolVsSam |> 
  pull(region) |> 
  factor()

D.1_ISABolVsSam_1 <- 
  multipatt(D.1_DIVBolVsSam, D.1_MetadataBolVsSam_1, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(D.1_ISABolVsSam_1)
```

Bolinao vs. Zambales.

```{r}
D.1_MetadataBolVsZam <- 
  A.1_Metadata_2 |> 
  filter(Species == "T. gigas",
         region %in% c("Bolinao", "Zambales")) |> 
  arrange(CorrectID)

D.1_DIVBolVsZam <- 
  C.1_DIVs_3 |> 
  filter(CorrectID %in% (D.1_MetadataBolVsZam |> pull(CorrectID))) |> 
  pivot_wider(names_from = "DIVs",
              values_from = "Count") |> 
  arrange(CorrectID) |> 
  column_to_rownames("CorrectID")

D.1_MetadataBolVsZam_1 <- 
  D.1_MetadataBolVsZam |> 
  pull(region) |> 
  factor()

D.1_ISABolVsZam_1 <- 
  multipatt(D.1_DIVBolVsZam, D.1_MetadataBolVsZam_1, 
                    control = how(nperm=999),
            duleg = TRUE)

summary(D.1_ISABolVsZam_1)
```
