---
title: "Within T. gigas"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load(vegan,
               tidyverse,
               janitor,
               rstatix,
               ggh4x,
               pairwiseAdonis,
               viridis,
               tidytext,
               ggthemes,
               ggtree,
               ape)
```

# Community

Load data.

```{r}
# ITS2 community
A.1_ITS2 <- read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

A.1_ITS2_1 <- 
  A.1_ITS2 |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  select(!c(1, 3:39))

# Metadata
A.1_Metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx")

A.1_Metadata_1 <- 
  A.1_Metadata |> 
  rename(sample_name = SymportalID)

A.1_Metadata_2 <- # full metadata
  A.1_Metadata_1 |> 
  filter(sample_name %in% (A.1_ITS2_1 |> pull(sample_name))) |> 
  arrange(CorrectID)

A.1_Metadata_3 <- # for filtering
  A.1_Metadata_2 |> 
  filter(sample_name %in% (A.1_ITS2_1 |> pull(sample_name))) |> 
  select(sample_name, CorrectID, Species, region) |> 
  arrange(CorrectID)

A.1_Metadata_4 <- # for filtering dbRDA
  A.1_Metadata_2 |> 
  filter(sample_name %in% (A.1_ITS2_1 |> pull(sample_name))) |> 
  filter(Species == "T. gigas")

# Merge
A.1_ITS2_2 <- 
  A.1_Metadata_3 |> 
  left_join(A.1_ITS2_1) |> 
  filter(Species == "T. gigas")
```

NMDS.

```{r}
B.1_ITS2 <- 
  A.1_ITS2_2 |> 
  select(!c(1,3,4)) |> 
  arrange(CorrectID)

B.1_ITS2_1 <- # turn into a matrix
  B.1_ITS2 |> 
  column_to_rownames("CorrectID") |> 
  sqrt()

B.1_ITS2_2 <- # turn into a distance
  B.1_ITS2_1 |> 
  vegdist(method = "bray")

## Do NMDS
B.1_ITS2NMDS <- 
  B.1_ITS2_2 |> 
  metaMDS(k = 2)

B.1_ITS2NMDS_1 <- # tidy NMDS data
  as.data.frame(scores(B.1_ITS2NMDS)) |> 
  rownames_to_column("CorrectID")

B.1_ITS2NMDS_2 <- # add metadata
  A.1_Metadata_3 |> 
  left_join(B.1_ITS2NMDS_1) |> 
  drop_na()
```

Plot.

```{r}
B.2_ITS2NMDS <- 
  B.1_ITS2NMDS_2 |> 
  ggplot(aes(x = NMDS1, y = NMDS2)) +
  geom_hline(yintercept = 0,
             color = 'darkgray',
             linetype = 'dashed',
             linewidth = 0.25) +
  geom_vline(xintercept = 0,
             color = 'darkgray',
             linetype = 'dashed',
             linewidth = 0.25) +
  geom_point(aes(color = region),
             size = 3) +
  labs(x = "NMDS1",
       y = "NMDS2",
       color = "Region") +
  scale_color_manual(values = 
                       c("Alaminos"="#E41A1C",
                         "Antique"="#377EB8",
                         "ARNP"="#4DAF4A",
                         "Batangas"="#984EA3",
                         "Bohol"="#FF7F00",
                         "Bolinao"="#FFFF33",
                         "Calaguas"="#A65628",
                         "Camiguin"="#F781BF",
                         "Palawan"="#999999",
                         "Samar"="#66C2A5",
                         "Zambales"="#FC8D62")) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.line = element_line(),panel.background    = element_rect(fill = "transparent", colour = NA),
        plot.background     = element_rect(fill = "transparent", colour = NA),
        legend.background   = element_rect(fill = "transparent", colour = NA),
        legend.box.background = element_rect(fill = "transparent", colour = NA),
        legend.position = "right",
        strip.text = element_text(face = "italic")) +
  coord_equal()

# Save
B.2_ITS2NMDS |> 
  ggsave(filename = "NMDS - Within Tgig - With Legend.svg", 
         dpi = 900,
         width = 17, 
         height = 8,
         unit = "cm")
```

PERMANOVA.

```{r}
B.3_PERMANOVA <- 
  adonis2(B.1_ITS2_2 ~ region, 
          A.1_Metadata_4,
          permutations = 9999)

B.3_PERMANOVAPwise <- # Pairwise PERMANOVA
  pairwise.adonis2(B.1_ITS2_2 ~ region,
                   data = A.1_Metadata_4,
                   nperm = 9999)

# Script to tidy PERMANOVA results
Z.fun_PPERMANOVACleaner <- 
  function(pmanova_res){
    a <- 
      pmanova_res[-1] |> 
      bind_rows(.id = 'groups') |> 
      drop_na() %>%
      .$`Pr(>F)` |> 
      p.adjust(method = 'BH') 
    
    b <- 
      unique(pmanova_res[-1] |> 
               bind_rows(.id = 'groups'))$groups |> 
      unique()
      
    d <- 
       a %>% 
        data.frame(b,.) |> 
        rename('groups' = 1,
               `P-adjusted` = 2) %>% 
        left_join(pmanova_res[-1] |> 
                    bind_rows(.id = 'groups'), 
                  ., 
                  by = 'groups') |> 
        mutate(`P-adjusted` = case_when(is.na(`Pr(>F)`) ~ NA,
                                        .default = `P-adjusted`)) |> 
        mutate(Parameter = rep(c('Species', 'Residual', 'Total'), length.out = n())) |> 
        relocate(last_col(), .after = 1)
    
    return(d)
  }

B.3_PERMANOVAPwise_1 <- # tidy
  B.3_PERMANOVAPwise |> 
  Z.fun_PPERMANOVACleaner()

B.3_PERMANOVAPwise_1 |> 
  write_csv("SourceVsTranplants.csv")
```

PERMDISP.

```{r}
B.1_PERMDISP <- 
  betadisper(B.1_ITS2_2, 
             group = (A.1_Metadata_4 |>
                        filter(CorrectID %in% (A.1_ITS2_2 |>
                                                 pull(CorrectID))))[["region"]]) |> 
  anova()
```

# ITS2 type profile

```{r}
B.1_ITS2TypeProf <- 
  read_tsv("../raw_data/20250829T091611/its2_type_profiles/4_first_analysis_20250829T091611.profiles.absolute.abund_and_meta.txt")

B.1_ITS2TypeProf_1 <- # tidy
  B.1_ITS2TypeProf |> 
  slice(-(1:5)) %>% # Remove first 5 rows (unnecessary rows)
  row_to_names(1) |>  # Move first row to column header
  rename(sample_name = 2) |>  # Name 2nd column
  select(-1) |>  # Remove first column (unneeded)
  drop_na() |> 
  type_convert() 

B.1_ITS2TypeProf_2 <- 
  A.1_Metadata_4 |> 
  select(sample_name, CorrectID, Species, region) |> 
  left_join(B.1_ITS2TypeProf_1) |> 
  select(!sample_name)

B.1_ITS2TypeProf_3 <- 
  B.1_ITS2TypeProf_2 |> 
  pivot_longer(!1:3) |> 
  filter(value > 0)

B.1_ITS2TypeProf_4 <- 
  B.1_ITS2TypeProf_3 |> 
  distinct(name) |> 
  pull(name)
```

Arrange based on ITS2 type profile clustering.

```{r}
# ITS2 type profile
## Cladocopium
# read distance matrix
B.1_Claodcopium <- 
  read_tsv("../raw_data/20250829T091611/between_profile_distances/C/20250829T091611_unifrac_profile_distances_C_sqrt.dist",
           col_names = F) |> 
  select(!2)

B.1_ClaodcopiumNames <- 
  B.1_Claodcopium |> 
  pull(X1)

B.1_ClaodcopiumNames <- 
  c("ITS2TypeProf", B.1_ClaodcopiumNames)

B.1_Claodcopium <- 
  B.1_Claodcopium |> 
  set_names(B.1_ClaodcopiumNames)

B.1_ClaodcopiumNames2 <- 
  B.1_Claodcopium |> 
  filter(ITS2TypeProf %in% B.1_ITS2TypeProf_4) |> 
  pull(ITS2TypeProf)

B.1_ClaodcopiumNames2 <- 
  c("ITS2TypeProf", B.1_ClaodcopiumNames2)

B.1_Claodcopium <- 
  B.1_Claodcopium |> 
  filter(ITS2TypeProf %in% B.1_ClaodcopiumNames2) |> 
  select(all_of(B.1_ClaodcopiumNames2)) |> 
  column_to_rownames("ITS2TypeProf")

# distance object
B.1_Claodcopium_2 <- as.dist(B.1_Claodcopium)

# UPGMA
B.1_Claodcopium_3 <- 
  hclust(B.1_Claodcopium_2,
         method = "average") |> 
  as.phylo()

# ggtree plot
B.1_Claodcopium_4 <- 
  B.1_Claodcopium_3 |> 
  ggtree() +
  coord_cartesian(clip = "off")+
  xlim(0, max(B.1_Claodcopium_3$edge.length) + 0.01)

# Save
B.1_Claodcopium_4 |> 
  ggsave(filename = "Tree - Within Tgig - C.svg", 
         dpi = 900,
         width = 2, 
         height = 3.25,
         unit = "cm")

B.1_Claodcopium_5 <- # tip order
  B.1_Claodcopium_4$data |>
  dplyr::filter(isTip) |>
  dplyr::arrange(y) |>
  dplyr::pull(label)


## Durusdinium
# read distance matrix
B.1_Durusdinium <- 
  read_tsv("../raw_data/20250829T091611/between_profile_distances/D/20250829T091611_unifrac_profile_distances_D_sqrt.dist",
           col_names = F) |> 
  select(!2)

B.1_DurusdiniumNames <- 
  B.1_Durusdinium |> 
  pull(X1)

B.1_DurusdiniumNames <- 
  c("ITS2TypeProf", B.1_DurusdiniumNames)

B.1_Durusdinium <- 
  B.1_Durusdinium |> 
  set_names(B.1_DurusdiniumNames)

B.1_DurusdiniumNames2 <- 
  B.1_Durusdinium |> 
  filter(ITS2TypeProf %in% B.1_ITS2TypeProf_4) |> 
  pull(ITS2TypeProf)

B.1_DurusdiniumNames2 <- 
  c("ITS2TypeProf", B.1_DurusdiniumNames2)

B.1_Durusdinium <- 
  B.1_Durusdinium |> 
  filter(ITS2TypeProf %in% B.1_DurusdiniumNames2) |> 
  select(all_of(B.1_DurusdiniumNames2)) |> 
  column_to_rownames("ITS2TypeProf")

# distance object
B.1_Durusdinium_2 <- as.dist(B.1_Durusdinium)

# UPGMA
B.1_Durusdinium_3 <- 
  hclust(B.1_Durusdinium_2,
         method = "average") |> 
  as.phylo()

# ggtree plot
B.1_Durusdinium_4 <- 
  B.1_Durusdinium_3 |> 
  ggtree() +
  coord_cartesian(clip = "off")+
  xlim(0, max(B.1_Durusdinium_3$edge.length) + 0.01)

# Save
B.1_Durusdinium_4 |> 
  ggsave(filename = "Tree - Within Tgig - D.svg", 
         dpi = 900,
         width = 4, 
         height = 8,
         unit = "cm")

B.1_Durusdinium_5 <- # tip order
  B.1_Durusdinium_4$data |>
  dplyr::filter(isTip) |>
  dplyr::arrange(y) |>
  dplyr::pull(label)

# Arrangement of ITS2 type profiles
B.1_ITS2TypeProfNames <- 
  c(B.1_Claodcopium_5, B.1_Durusdinium_5, "A3")
```

Plot.

```{r}
# Plot
B.2_ITS2TypeProf <- 
  B.1_ITS2TypeProf_3 |> 
  group_by(CorrectID) |>
  mutate(RelAbun = value/sum(value)) |> 
  ungroup() |>
  select(CorrectID, name, RelAbun) |> 
  pivot_wider(names_from = name,
              values_from = RelAbun,
              values_fill = 0) |> 
  pivot_longer(!1,
               names_to = "ITS2TypeProf",
               values_to = "RelAbun") |> 
  mutate(RelAbun = sqrt(RelAbun),
         Clade = str_extract(ITS2TypeProf, "[A-Z]")) |>
  left_join(A.1_Metadata_4) |>
  mutate(ITS2TypeProf = factor(ITS2TypeProf, levels = B.1_ITS2TypeProfNames)) |> 
  ggplot(aes(x = CorrectID,
             y = ITS2TypeProf,
             fill = RelAbun)) +
  geom_raster() +
  facet_grid(cols = vars(region),
             rows = vars(Clade),
             space = "free",
             scale = "free") +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        strip.background = element_blank(),
        panel.spacing = unit(0.2,'lines'),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size = 7, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom",) +
  scale_fill_viridis(option = "cividis")

# Save
B.2_ITS2TypeProf |> 
  ggsave(filename = "Tile - Within Tgig - With Legend.svg", 
         dpi = 900,
         width = 16, 
         height = 7,
         unit = "cm")
```

# PERMANOVA with covariates

```{r}
C.1_PERMANOVAApproxShellLength <- 
  adonis2(B.1_ITS2_2 ~ ApproxShellLength, 
          A.1_Metadata_4,
          permutations = 9999)

C.1_PERMANOVAApproxAge <- 
  adonis2(B.1_ITS2_2 ~ ApproxAge, 
          A.1_Metadata_4,
          permutations = 9999)

C.1_PERMANOVAAgeTransplanted <- 
  adonis2(B.1_ITS2_2 ~ AgeTransplanted, 
          A.1_Metadata_4,
          permutations = 9999)
```

# Camiguin

```{r}
D.1_Metadata <- 
  A.1_Metadata_3 |> 
  filter(region == "Camiguin")

D.1_ITS2TypeProf <- 
  D.1_Metadata |> 
  left_join(B.1_ITS2TypeProf_1)
```

Plot.

```{r}
# Tidy
D.1_ITS2TypeProf_1 <- 
  D.1_ITS2TypeProf |> 
  pivot_longer(!1:4) |> 
  group_by(CorrectID) |>
  mutate(RelAbun = value/sum(value)) |> 
  ungroup() |>
  select(CorrectID, name, RelAbun) |> 
  filter(RelAbun > 0) |> 
  pivot_wider(names_from = name,
              values_from = RelAbun,
              values_fill = 0) |> 
  pivot_longer(!1,
               names_to = "ITS2TypeProf",
               values_to = "RelAbun") |> 
  mutate(RelAbun = sqrt(RelAbun),
         Clade = str_extract(ITS2TypeProf, "[A-Z]")) |>
  left_join(D.1_Metadata)

# Plot
D.2_ITS2TypeProf <- 
  D.1_ITS2TypeProf_1 |> 
  ggplot(aes(x = CorrectID,
             y = ITS2TypeProf,
             fill = RelAbun)) +
  geom_tile(color = "white",
            linewidth = 0.1) +
  facet_grid(cols = vars(Species),
             rows = vars(Clade),
             space = "free",
             scale = "free") +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        strip.background = element_blank(),
        panel.spacing = unit(0.2,'lines'),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 7),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  scale_fill_viridis(option = "rocket")
  
# Save
D.2_ITS2TypeProf |> 
  ggsave(filename = "Tile - Within Camiguin - With Legend.svg", 
         dpi = 900,
         width = 17, 
         height = 8,
         unit = "cm")
```
