---
title: "Photosymbs in GCs"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load(tidyverse,
               janitor,
               vegan,
               cowplot,
               pairwiseAdonis,
               svglite,
               viridis,
               multcompView,
               scico,
               ggbeeswarm,
               ggupset,
               rstatix,
               ape,
               ggtree)
```

# Genus

Tidy data.

```{r}
B.1_ITS2 <- # clean its2 data with correct id info
  A.1_ITS2_2

B.1_ITS2_1 <- # pivot longer
  B.1_ITS2 |> 
  select(!c(1,3,4)) |> 
  pivot_longer(!1,
               names_to = "ITS2",
               values_to = "Count")

B.1_ITS2_2 <- # merge with metadata
  B.1_ITS2_1 |> 
  left_join(A.1_Metadata_1)

B.1_ITS2_3 <- # genus info
  B.1_ITS2_2 |> 
  # Detect genus
  mutate(Genus = case_when(str_detect(ITS2, "A") ~ "Symbiodinium",
                           str_detect(ITS2, "C") ~ "Cladocopium",
                           str_detect(ITS2, "D") ~ "Durusdinium",
                           str_detect(ITS2, "G") ~ "Gerakladium",
                           .default = NA)) |> 
  drop_na() |> 
  group_by(CorrectID, Genus, Species, region) |> 
  summarise(Total = sum(Count)) |> 
  ungroup() |> 
  arrange(tolower(region)) %>%     # sort the data frame by region, then ID
  mutate(CorrectID = factor(CorrectID, levels = unique(CorrectID)))
```

Plot.

```{r}
B.2_ITS2Genus <- 
  B.1_ITS2_3 |> 
  ggplot(aes(x = CorrectID,
             y = Total,
             fill = fct_relevel(Genus, c("Symbiodinium",
                                         "Cladocopium",
                                         "Durusdinium",
                                         "Gerakladium")))) +
  geom_col(position = "fill",
           width = 1.15,
           color = NA,
           linewidth = 0) +
  facet_grid(cols = vars(fct_relevel(Species,
                                     c("H. hippopus",
                                      "H. porcellanus",
                                      "T. crocea",
                                      "T. squamosa",
                                      "T. noae",
                                      "T. maxima",
                                      "T. derasa",
                                      "T. gigas"))),
             space = "free",
             scale = "free") +
  scale_fill_manual(values = 
                      c("Symbiodinium"= "#746FB4",
                        "Cladocopium" = "#209E77",
                        "Durusdinium" = "#E7AB06",
                        "Gerakladium" = "#A9A9A9")) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        strip.background = element_blank(),
        panel.spacing = unit(0.1,'lines'),
        axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        axis.line.y = element_line(),
        axis.ticks.y = element_line())
```

Save.

```{r}
(B.2_ITS2Genus +
    theme(legend.position = "none")) |> 
  ggsave(filename = "Barplot - Genus.svg", 
         dpi = 900,
         width = 14.75, 
         height = 2,
         unit = "cm")

B.2_ITS2Genus |> 
  ggsave(filename = "Barplot - Genus - With Legend.svg", 
         dpi = 900,
         width = 20, 
         height = 10,
         unit = "cm")
```

# Region labels

Tidy data.

```{r}
E.1_Metadata <- # filter metadata to include samples with ITS2 seqs
  A.1_Metadata_1 |> 
  filter(CorrectID %in% (A.1_ITS2_2 |> 
                           pull(CorrectID)))

E.1_Metadata_1 <- # add counts which will be used to do barplots/labels
  E.1_Metadata |> 
  mutate(Count = 1)

E.1_Metadata_2 <- # arrange
  E.1_Metadata_1 |> 
  arrange(CorrectID)
```

Plot.

```{r}
E.2_Region <- 
  E.1_Metadata_2 |> 
  ggplot(aes(x = fct_reorder(CorrectID, region),
             y = Count,
             fill = region)) +
  geom_col(width = 1.15) +
  facet_grid(cols = vars(fct_relevel(Species,
                                    c("H. hippopus",
                                      "H. porcellanus",
                                      "T. crocea",
                                      "T. squamosa",
                                      "T. noae",
                                      "T. maxima",
                                      "T. derasa",
                                      "T. gigas"))),
             scales = "free",
             space = "free") +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  scale_fill_manual(values = c("Alaminos"="#E41A1C",
                               "Antique"="#377EB8",
                               "ARNP"="#4DAF4A",
                               "Batangas"="#984EA3",
                               "Bohol"="#FF7F00",
                               "Bolinao"="#FFFF33",
                               "Calaguas"="#A65628",
                               "Camiguin"="#F781BF",
                               "Palawan"="#999999",
                               "Samar"="#66C2A5",
                               "Zambales"="#FC8D62")) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        panel.spacing = unit(0.1,'lines'),
        axis.title = element_blank(),
        legend.position = "top") +
  guides(fill = guide_legend(nrow = 1))
```

Save plot.

```{r}
(E.2_Region +
    theme(legend.position = "none",
          axis.text = element_blank()))  |> 
  ggsave(filename = "Labels - Regions - No legend.svg", 
         dpi = 900,
         width = 17.5, 
         height = 4,
         unit = "cm")

E.2_Region |> 
  ggsave(filename = "Labels - Regions.svg", 
         dpi = 900,
         width = 25, 
         height = 4,
         unit = "cm")
```

# ITS2 type profile

```{r}
F.1_ITS2TypeProf <- 
  read_tsv("../raw_data/20250829T091611/its2_type_profiles/4_first_analysis_20250829T091611.profiles.absolute.abund_and_meta.txt")

F.1_ITS2TypeProf_1 <- # tidy
  F.1_ITS2TypeProf |> 
  slice(-(1:5)) %>% # Remove first 5 rows (unnecessary rows)
  row_to_names(1) |>  # Move first row to column header
  rename(sample_name = 2) |>  # Name 2nd column
  select(-1) |>  # Remove first column (unneeded)
  drop_na() |> 
  type_convert() 

F.1_ITS2TypeProf_2 <- 
  E.1_Metadata_1 |> 
  select(sample_name, CorrectID, Species, region) |> 
  left_join(F.1_ITS2TypeProf_1) |> 
  select(!sample_name)

F.1_ITS2TypeProf_3 <- 
  read_csv("../raw_data/Edited ITS2 type profile naming.csv") |> 
  rename(name = 1)

F.1_ITS2TypeProf_4 <- 
  F.1_ITS2TypeProf_2 |> 
  pivot_longer(!1:3) |> 
  filter(value > 0) |> 
  left_join(F.1_ITS2TypeProf_3)

F.1_ITS2TypeProf_5 <- # ITS2 type profiles
  F.1_ITS2TypeProf_4 |> 
  distinct(name) |> 
  pull(name)
```

Initial plot.

```{r}
F.2_ITS2TypeProf <- 
  F.1_ITS2TypeProf_4 |> 
  group_by(CorrectID) |>
  mutate(RelAbun = value/sum(value)) |> 
  ungroup() |>
  select(CorrectID, Edited, RelAbun) |> 
  pivot_wider(names_from = Edited,
              values_from = RelAbun,
              values_fill = 0) |> 
  pivot_longer(!1,
               names_to = "ITS2TypeProf",
               values_to = "RelAbun") |> 
  mutate(RelAbun = sqrt(RelAbun),
         Clade = str_extract(ITS2TypeProf, "[A-Z]")) |>
  left_join(E.1_Metadata_1) |> 
  ggplot(aes(x = CorrectID,
             y = ITS2TypeProf,
             fill = RelAbun)) +
  geom_raster() +
  facet_grid(cols = vars(Species),
             rows = vars(Clade),
             space = "free",
             scale = "free") +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        strip.background = element_blank(),
        panel.spacing = unit(0.2,'lines'),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 7, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  scale_fill_viridis(option = "cividis")
```

Add Unifrac data.

*Cladocopium*.

```{r}
## Cladocopium
# read distance matrix
G.1_Cladocopium <- 
  read_tsv("../raw_data/20250829T091611/between_profile_distances/C/20250829T091611_unifrac_profile_distances_C_no_sqrt.dist",
           col_names = F) |> 
  select(!2)

G.1_CladocopiumNames <- 
  G.1_Cladocopium |> 
  pull(X1)

G.1_CladocopiumNames <- 
  c("ITS2TypeProf", G.1_CladocopiumNames)

G.1_Cladocopium <- 
  G.1_Cladocopium |> 
  set_names(G.1_CladocopiumNames)

G.1_CladocopiumNames2 <- 
  G.1_Cladocopium |> 
  filter(ITS2TypeProf %in% F.1_ITS2TypeProf_5) |> 
  pull(ITS2TypeProf)

G.1_CladocopiumNames2 <- 
  c("ITS2TypeProf", G.1_CladocopiumNames2)

G.1_Cladocopium <- 
  G.1_Cladocopium |> 
  filter(ITS2TypeProf %in% G.1_CladocopiumNames2) |> 
  select(all_of(G.1_CladocopiumNames2)) |> 
  column_to_rownames("ITS2TypeProf")

# distance object
G.1_Cladocopium_2 <- as.dist(G.1_Cladocopium)

# UPGMA
G.1_Cladocopium_3 <- 
  hclust(G.1_Cladocopium_2,
         method = "average") |> 
  as.phylo()

# ggtree plot
G.1_Cladocopium_4 <- 
  G.1_Cladocopium_3 |> 
  ggtree() +
  geom_tiplab(size = 3) +
  coord_cartesian(clip = "off") +
  xlim(0, max(G.1_Cladocopium_3$edge.length) + 0.01)

# Save
G.1_Cladocopium_4 |> 
  ggsave(filename = "Tree - Within Tgig - C.svg", 
         dpi = 900,
         width = 6, 
         height = 5,
         unit = "cm")

G.1_Cladocopium_5 <- # tip order
  G.1_Cladocopium_4$data |>
  dplyr::filter(isTip) |>
  dplyr::arrange(y) |>
  dplyr::pull(label)
```

*Durusdinium*.

```{r}
## Durusdinium
# read distance matrix
G.1_Durusdinium <- 
  read_tsv("../raw_data/20250829T091611/between_profile_distances/D/20250829T091611_unifrac_profile_distances_D_no_sqrt.dist",
           col_names = F) |> 
  select(!2)

G.1_DurusdiniumNames <- 
  G.1_Durusdinium |> 
  pull(X1)

G.1_DurusdiniumNames <- 
  c("ITS2TypeProf", G.1_DurusdiniumNames)

G.1_Durusdinium <- 
  G.1_Durusdinium |> 
  set_names(G.1_DurusdiniumNames)

G.1_DurusdiniumNames2 <- 
  G.1_Durusdinium |> 
  filter(ITS2TypeProf %in% F.1_ITS2TypeProf_5) |> 
  pull(ITS2TypeProf)

G.1_DurusdiniumNames2 <- 
  c("ITS2TypeProf", G.1_DurusdiniumNames2)

G.1_Durusdinium <- 
  G.1_Durusdinium |> 
  filter(ITS2TypeProf %in% G.1_DurusdiniumNames2) |> 
  select(all_of(G.1_DurusdiniumNames2)) |> 
  column_to_rownames("ITS2TypeProf")

# distance object
G.1_Durusdinium_2 <- as.dist(G.1_Durusdinium)

# UPGMA
G.1_Durusdinium_3 <- 
  hclust(G.1_Durusdinium_2,
         method = "average") |> 
  as.phylo()

# ggtree plot
G.1_Durusdinium_4 <- 
  G.1_Durusdinium_3 |> 
  ggtree() +
  geom_tiplab(size = 3) +
  coord_cartesian(clip = "off") +
  xlim(0, max(G.1_Durusdinium_3$edge.length) + 0.01)

# Save
G.1_Durusdinium_4 |> 
  ggsave(filename = "Tree - Within Tgig - D.svg", 
         dpi = 900,
         width = 6, 
         height = 5,
         unit = "cm")

G.1_Durusdinium_5 <- # tip order
  G.1_Durusdinium_4$data |>
  dplyr::filter(isTip) |>
  dplyr::arrange(y) |>
  dplyr::pull(label)
```

*Symbiodinium*.

```{r}
## Symbiodinium
# read distance matrix
G.1_Symbiodinium <- 
  read_tsv("../raw_data/20250829T091611/between_profile_distances/A/20250829T091611_unifrac_profile_distances_A_no_sqrt.dist",
           col_names = F) |> 
  select(!2)

G.1_SymbiodiniumNames <- 
  G.1_Symbiodinium |> 
  pull(X1)

G.1_SymbiodiniumNames <- 
  c("ITS2TypeProf", G.1_SymbiodiniumNames)

G.1_Symbiodinium <- 
  G.1_Symbiodinium |> 
  set_names(G.1_SymbiodiniumNames)

G.1_SymbiodiniumNames2 <- 
  G.1_Symbiodinium |> 
  filter(ITS2TypeProf %in% F.1_ITS2TypeProf_5) |> 
  pull(ITS2TypeProf)

G.1_SymbiodiniumNames2 <- 
  c("ITS2TypeProf", G.1_SymbiodiniumNames2)

G.1_Symbiodinium <- 
  G.1_Symbiodinium |> 
  filter(ITS2TypeProf %in% G.1_SymbiodiniumNames2) |> 
  select(all_of(G.1_SymbiodiniumNames2)) |> 
  column_to_rownames("ITS2TypeProf")

# distance object
G.1_Symbiodinium_2 <- as.dist(G.1_Symbiodinium)

# UPGMA
G.1_Symbiodinium_3 <- 
  hclust(G.1_Symbiodinium_2,
         method = "average") |> 
  as.phylo()

# ggtree plot
G.1_Symbiodinium_4 <- 
  G.1_Symbiodinium_3 |> 
  ggtree() +
  geom_tiplab(size = 3) +
  coord_cartesian(clip = "off") +
  xlim(0, max(G.1_Symbiodinium_3$edge.length) + 0.01)

# Save
G.1_Symbiodinium_4 |> 
  ggsave(filename = "Tree - Within Tgig - A.svg", 
         dpi = 900,
         width = 6, 
         height = 5,
         unit = "cm")

G.1_Symbiodinium_5 <- # tip order
  G.1_Symbiodinium_4$data |>
  dplyr::filter(isTip) |>
  dplyr::arrange(y) |>
  dplyr::pull(label)
```

Arrange y-axis based on the trees.

```{r}
H.1_ITS2TypeProfNames <- 
  c(G.1_Cladocopium_5, G.1_Durusdinium_5, G.1_Symbiodinium_5) %>%
  tibble("name" = .) |> 
  left_join(F.1_ITS2TypeProf_3) |> 
  pull(Edited)
```

Arrange plot.

```{r}
H.2_ITS2TypeProf <- 
  F.1_ITS2TypeProf_4 |> 
  group_by(CorrectID) |>
  mutate(RelAbun = value/sum(value)) |> 
  ungroup() |>
  select(CorrectID, Edited, RelAbun) |> 
  pivot_wider(names_from = Edited,
              values_from = RelAbun,
              values_fill = 0) |> 
  pivot_longer(!1,
               names_to = "ITS2TypeProf",
               values_to = "RelAbun") |> 
  mutate(RelAbun = sqrt(RelAbun),
         Clade = str_extract(ITS2TypeProf, "[A-Z]")) |>
  left_join(E.1_Metadata_1) |>
  mutate(ITS2TypeProf = factor(ITS2TypeProf, levels = H.1_ITS2TypeProfNames)) |> 
  ungroup() |> 
  arrange(tolower(region)) %>%     # sort the data frame by region, then ID
  mutate(CorrectID = factor(CorrectID, levels = unique(CorrectID))) |> 
  ggplot(aes(x = CorrectID,
             y = ITS2TypeProf,
             fill = RelAbun)) +
  geom_raster() +
  facet_grid(cols = vars(fct_relevel(Species, c("H. hippopus",
                                      "H. porcellanus",
                                      "T. crocea",
                                      "T. squamosa",
                                      "T. noae",
                                      "T. maxima",
                                      "T. derasa",
                                      "T. gigas"))),
             rows = vars(Clade),
             space = "free",
             scale = "free") +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        strip.background = element_blank(),
        panel.spacing = unit(0.2,'lines'),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 7, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  scale_fill_viridis(option = "cividis")

H.2_ITS2TypeProf |> 
  ggsave(filename = "ITS2 type profile - By species.svg", 
         dpi = 900,
         width = 15, 
         height = 11,
         unit = "cm")
```

# Lineages

## Species

```{r}
I.1_ITS2TypeProf <- 
  F.1_ITS2TypeProf_4

I.1_ITS2TypeProf_1 <- # get the dominant DIV
  I.1_ITS2TypeProf |> 
  separate(Edited, into = "lineage", sep = "-", extra = "drop", remove = FALSE)

I.1_ITS2TypeProf_2 <- # get total per sample per lineage per species
  I.1_ITS2TypeProf_1 |> 
  mutate(lineage = if_else(lineage == "C66/C93a", "C93a/C66", lineage)) |> # consolidate
  mutate(lineage = if_else(lineage == "D4/D5/D9b", "D5/D4/D9b", lineage))	|> # consolidate
  group_by(CorrectID, Species, region, lineage) |> 
  summarise(Total = sum(value)) |> 
  ungroup() |> 
  group_by(CorrectID) |> 
  mutate(RelativeAbundance = Total/sum(Total)) |> 
  ungroup()

I.1_ITS2TypeProf_3 <- # get dominant lineages per sample (>=0.6)
  I.1_ITS2TypeProf_2 |> 
  mutate(IsDominant = if_else(RelativeAbundance >= 0.8, 1, 0)) |> 
  group_by(Species, lineage) |> 
  summarise(Total = sum(IsDominant)) |> 
  ungroup()

I.1_ITS2TypeProf_4 <- # no. of samples per species
  E.1_Metadata |> 
  count(Species)

I.1_ITS2TypeProf_5 <- 
  I.1_ITS2TypeProf_3 |> 
  left_join(I.1_ITS2TypeProf_4) |> 
  mutate(Prop = Total/n *100)
```

## Region

```{r}
J.1_ITS2TypeProf <- # get total per sample per lineage per region
  I.1_ITS2TypeProf_2 |> 
  mutate(IsDominant = if_else(RelativeAbundance >= 0.6, 1, 0)) |> 
  group_by(region, lineage) |> 
  summarise(Total = sum(IsDominant)) |> 
  ungroup()

J.1_ITS2TypeProf_1 <- # get dominant lineages per sample (>=0.6)
  E.1_Metadata |> 
  count(region)

J.1_ITS2TypeProf_2 <- 
  J.1_ITS2TypeProf |> 
  left_join(J.1_ITS2TypeProf_1) |> 
  mutate(Prop = Total/n *100)
```
