---
title: "Photosymbionts of Philippine GCs"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load(tidyverse,
               janitor,
               vegan,
               cowplot,
               viridis,
               rstatix,
               ggbeeswarm)
```

# Box plot

Load metadata.

```{r}
A.1_Metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx")

A.1_Metadata_1 <- 
  A.1_Metadata |> 
  select(c(1, 2, 3, 12)) |> 
  rename(sample_name = 1)
```

Load Symportal analysis results.

```{r}
A.1_ITS2 <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

A.1_ITS2_1 <- 
  A.1_ITS2 |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  left_join(A.1_Metadata_1) |> 
  relocate(sample_name, CorrectID, Species, region)

A.1_ITS2_2 <- # tidy
  A.1_ITS2_1 |> 
  select(CorrectID, Species, region, 43:503)

A.1_ITS2_3 <- # pivot longer, get capital letter
  A.1_ITS2_2 |> 
  pivot_longer(!c(1:3),
               names_to = "ITS2Sequence",
               values_to = "Count") |> 
  mutate(Clade = str_extract(ITS2Sequence, "[A-Z]")) |> 
  filter(Clade != "G")

A.1_ITS2_4 <- # get total counts per genus per sample
  A.1_ITS2_3 |> 
  group_by(CorrectID, Clade) |> 
  summarise(Total = sum(Count)) |> 
  ungroup()

A.1_ITS2_5 <- # get relative abundance per genus per sample 
  A.1_ITS2_4 |> 
  group_by(CorrectID) |> 
  mutate(RelativeAbundance = Total/sum(Total)*100) |> 
  ungroup()
```

Plot.

```{r}
A.2_ComparisonAmongGenus <- 
  A.1_ITS2_5 |> 
  ggplot(aes(x = Clade,
             y = RelativeAbundance,
             color = Clade,
             fill = Clade)) +
  stat_summary(fun = "mean",
               geom = "bar",
               alpha = 0.75,
               color = NA,
               width = 0.75) +
  stat_summary(fun = mean,
               fun.data = mean_se,
               geom = "linerange") +
  geom_jitter(width = 0.1, alpha = 0.25) +
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,100)) +
  theme_minimal() +
  coord_cartesian(clip = "off") +
  theme(panel.grid = element_blank(),
        axis.line.y = element_line(),
        axis.ticks.y = element_line(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none") +
  scale_fill_manual(values = 
                      c("A"= "#746FB4",
                        "C" = "#209E77",
                        "D" = "#E7AB06",
                        "G" = "#A9A9A9")) +
  scale_color_manual(values = 
                      c("A"= "#746FB4",
                        "C" = "#209E77",
                        "D" = "#E7AB06",
                        "G" = "#A9A9A9"))
```

Stats.

```{r}
A.3_ITS2 <- 
  A.1_ITS2_5 |> 
  friedman_test(RelativeAbundance ~ Clade | CorrectID)

A.3_ITS2Pair <- 
  A.1_ITS2_5 |> 
  pairwise_wilcox_test(
    RelativeAbundance ~ Clade, 
    paired = TRUE,
    p.adjust.method = "BH")
```

Save.

```{r}
A.2_ComparisonAmongGenus |> 
  ggsave(filename = "Relative abundance of genera.svg", 
         dpi = 900,
         width = 5, 
         height = 5.2,
         unit = "cm")
```

# Barplot

Tidy.

```{r}
B.1_ITS2 <- 
  A.1_ITS2_4 |> 
  group_by(Clade) |> 
  summarise(Count = sum(Total)) |> 
  ungroup() |> 
  mutate(RelAb = Count/sum(Count))
```

Plot.

```{r}
B.2_ITS2 <- 
  B.1_ITS2 |> 
  ggplot(aes(x = 1,
             y = RelAb,
             fill = Clade)) +
  geom_bar(position = "fill",
           stat = "identity") +
  theme_minimal() +
  scale_fill_manual(values = 
                      c("A"= "#746FB4",
                        "C" = "#209E77",
                        "D" = "#E7AB06",
                        "G" = "#A9A9A9")) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0,1),
                     labels = scales::label_percent()) +
  theme(panel.grid = element_blank(),
        axis.line.y = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_blank(),
        axis.text.x = element_blank(),
        legend.title = element_blank(),
        legend.position = "none")
```

Save.

```{r}
B.2_ITS2 |> 
  ggsave(filename = "Donut.svg", 
         dpi = 900,
         width = 3, 
         height = 8,
         unit = "cm")
```

# ITS2 seqs

Tidy.

```{r}
C.1_ITS2 <- # pivot longer
  A.1_ITS2_2 |> 
  pivot_longer(!c(1:3),
               names_to = "ITS2",
               values_to = "Count")

C.1_ITS2_1 <- # get total per sequence, per clade
  C.1_ITS2 |> 
  group_by(ITS2) |> 
  summarize(Count = sum(Count)) |> 
  mutate(Clade = str_extract(ITS2, "[A-Z]")) 
```

ITS2 sequences with \>1% relative abundance.

```{r}
 C.1_ITS2_2 <- 
   C.1_ITS2_1 |> 
   mutate(RelAbun = Count/sum(Count) *100)
 
 C.1_ITS2_3 <- 
   C.1_ITS2_2 |> 
   filter(RelAbun >= 1) |> 
   pull(ITS2)
```

*Symbiodinium.*

```{r}
C.1_A <- 
  C.1_ITS2_1 |> 
  filter(Clade == "A") |> 
  mutate(Group = if_else(ITS2 %in% C.1_ITS2_3, ITS2, "Others")) |> 
  group_by(Group) |> 
  summarise(Count = sum(Count))

C.2_A <- 
  C.1_A |> 
  ggplot(aes(x = "A",
             y = Count,
             fill = fct_relevel(Group, c("A3", "A6b", "Others")))) +
  geom_bar(stat = "identity",
           position = "fill") +
  scale_y_continuous(expand = c(0,0),
                     labels = scales::label_percent()) +
  theme_minimal() +
  scale_fill_manual(values = c("#2B284C","#746FB4", "darkgray")) +
  coord_cartesian(clip = "off") +
  theme(panel.grid = element_blank(),
        axis.line.y = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_blank(),
        axis.text.x = element_blank(),
        legend.title = element_blank())
```

*Cladocopium.*

```{r}
C.1_C <- 
  C.1_ITS2_1 |> 
  filter(Clade == "C") |> 
  mutate(Group = if_else(ITS2 %in% C.1_ITS2_3, ITS2, "Others")) |> 
  group_by(Group) |> 
  summarise(Count = sum(Count))

C.2_C <- 
  C.1_C |> 
  ggplot(aes(x = "A",
             y = Count,
             fill = fct_relevel(Group, c("C93a", "C1", "C66", "C1c", "Others")))) +
  geom_bar(stat = "identity",
           position = "fill") +
  scale_y_continuous(expand = c(0,0),
                     labels = scales::label_percent()) +
  scale_fill_manual(values = c("#0F4534","#146048","#209E77", "#B0EEDB", "darkgray")) +
  theme_minimal() +
  coord_cartesian(clip = "off") +
  theme(panel.grid = element_blank(),
        axis.line.y = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_blank(),
        axis.text.x = element_blank(),
        legend.title = element_blank())
```

*Durusdinium.*

```{r}
C.1_D <- 
  C.1_ITS2_1 |> 
  filter(Clade == "D") |> 
  mutate(Group = if_else(ITS2 %in% C.1_ITS2_3, ITS2, "Others")) |> 
  group_by(Group) |> 
  summarise(Count = sum(Count))

C.2_D <- 
  C.1_D |> 
  ggplot(aes(x = "A",
             y = Count,
             fill = fct_relevel(Group, c("D5", "D4", "D9b", "D1", "Others")))) +
  geom_bar(stat = "identity",
           position = "fill") +
  scale_y_continuous(expand = c(0,0),
                     labels = scales::label_percent()) +
  theme_minimal() +
  scale_fill_manual(values = c("#E67800","#E7AB06","#FAC738", "#FDEAB5", "darkgray")) +
  coord_cartesian(clip = "off") +
  theme(panel.grid = element_blank(),
        axis.line.y = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_blank(),
        axis.text.x = element_blank(),
        legend.title = element_blank())
```

Merge.

```{r}
C.2_Merged <- 
  plot_grid(C.2_A,
            C.2_C,
            C.2_D,
            nrow = 1)

C.2_Merged |> # save
  ggsave(filename = "Top ITS2 sequences.svg", 
         dpi = 900,
         width = 13.25, 
         height = 4.45,
         unit = "cm")
```

# RA of ITS2 sequence

Tidy.

```{r}
D.1_PresenceITS2 <- # presence
  A.1_ITS2_2 |>
  select(!c(2,3)) |> 
  pivot_longer(!1,
               names_to = "ITS2",
               values_to = "Count") |> 
  mutate(Presence = if_else(Count > 0, 1, 0)) |> # get presence data
  group_by(ITS2) |> # group then get total # of colonies with specific ITS2
  summarize(TotalPresence = sum(Presence)) |> 
  ungroup() |> 
  filter(ITS2 %in% C.1_ITS2_3) |> # filter ITS2 with >1% rel abun
  mutate(ProportionGCs = TotalPresence/165) |> # get proportion
  mutate(Clade = str_extract(ITS2, "[A-Z]"))  # get clade info
```

Plot.

```{r}
D.2_PresenceITS2 <- 
  D.1_PresenceITS2 |> 
  ggplot(aes(x = fct_reorder(ITS2, desc(ProportionGCs)),
             y = ProportionGCs,
             color = Clade,
             fill = Clade)) +
  geom_col() +
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,0.75),
                     breaks = seq(0, 0.75, by = 0.25),
                     labels = scales::label_percent()) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.line.y = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        axis.title = element_blank()) +
  scale_fill_manual(values = 
                      c("A"= "#746FB4",
                        "C" = "#209E77",
                        "D" = "#E7AB06",
                        "G" = "#A9A9A9")) +
  scale_color_manual(values = 
                      c("A"= "#746FB4",
                        "C" = "#209E77",
                        "D" = "#E7AB06",
                        "G" = "#A9A9A9"))
```

Save.

```{r}
D.2_PresenceITS2 |> # save
  ggsave(filename = "Presence and abundance.svg", 
         dpi = 900,
         width = 10.3, 
         height = 3.3,
         unit = "cm")
```

# ITS2 type profiles

Load data.

```{r}
E.1_ITS2TypeProf <- 
  read_tsv("../raw_data/20250829T091611/its2_type_profiles/4_first_analysis_20250829T091611.profiles.absolute.abund_and_meta.txt")

E.1_ITS2TypeProf_1 <- # tidy
  E.1_ITS2TypeProf |> 
  slice(-(1:5)) %>% # Remove first 5 rows (unnecessary rows)
  row_to_names(1) |>  # Move first row to column header
  rename(sample_name = 2) |>  # Name 2nd column
  select(-1) |>  # Remove first column (unneeded)
  drop_na() |> 
  type_convert() 

E.1_ITS2TypeProf_2 <- # merge with metadata, remove unneeded
  E.1_ITS2TypeProf_1 |> 
  left_join(A.1_Metadata_1) |> 
  filter(sample_name %in% (A.1_ITS2_1 |> pull(sample_name))) |> 
  relocate(1, 35:37)

E.1_ITS2TypeProf_3 <- # tidy
  E.1_ITS2TypeProf_2 |> 
  select(!c(1,3,4))

E.1_ITS2TypeProf_4 <- # manual edits
  E.1_ITS2TypeProf_3 |> edit()

E.1_ITS2TypeProf_5 <- #pivot 
  E.1_ITS2TypeProf_4 |> 
  pivot_longer(!CorrectID,
               names_to = "ITS2TypeProf",
               values_to = "Counts") |> 
  group_by(ITS2TypeProf) |> # get sum per ITS2 type profile
  summarize(Total = sum(Counts)) |> 
  mutate(Clade = str_extract(ITS2TypeProf, "[A-Z]")) |> 
  ungroup() |> 
  mutate(Denominator = sum(Total),
         RelativeAbundance = Total/Denominator)
```

Plot.

```{r}
E.2_ITS2TypeProf <- 
  E.1_ITS2TypeProf_5 |> 
  ggplot(aes(x = fct_reorder(ITS2TypeProf, desc(RelativeAbundance)),
             y = RelativeAbundance,
             fill = Clade)) +
  geom_col() +
  coord_cartesian(clip = "off") +
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,0.20),
                     labels = scales::label_percent()) +
  facet_grid(cols = vars(Clade),
             space = "free",
             scales = "free") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.line.y = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        axis.title = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.spacing = unit(0.1,'lines')) +
  scale_fill_manual(values = 
                      c("A"= "#746FB4",
                        "C" = "#209E77",
                        "D" = "#E7AB06",
                        "G" = "#A9A9A9"))
```

Save.

```{r}
E.2_ITS2TypeProf |> # save
  ggsave(filename = "ITS2 type profile.svg", 
         dpi = 900,
         width = 16, 
         height = 10,
         unit = "cm")
```
