---
title: "Sequencing stats"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load(tidyverse,
               janitor,
               vegan,
               cowplot,
               viridis)
```

# ITS2 results

Load metadata.

```{r}
A.1_Metadata <- 
  readxl::read_xlsx("../metadata/metadata_20251021.xlsx")

A.1_Metadata_1 <- 
  A.1_Metadata |> 
  select(c(1, 2, 3, 12)) |> 
  rename(sample_name = 1)
```

Load Symportal analysis results.

```{r}
A.1_ITS2 <- 
  read_tsv("../raw_data/20250829T091611/post_med_seqs/4_first_analysis_20250829T091611.seqs.absolute.abund_and_meta.txt")

A.1_ITS2_1 <- 
  A.1_ITS2 |> 
  drop_na()  |> 
  filter(post_med_unique != 0) |> 
  left_join(A.1_Metadata_1) |> 
  relocate(sample_name, CorrectID, Species, region)

A.1_ITS2_2 <- # tidy
  A.1_ITS2_1 |> 
  select(CorrectID, Species, region, 12, 13, 43:503)
```

Plot.

```{r}
A.2_ITS2 <- 
  A.1_ITS2_2 |> 
  mutate(raw_contigs = raw_contigs/10000,
         post_qc_absolute_seqs = post_qc_absolute_seqs/10000) |> 
  ggplot() +
  geom_col(aes(x = CorrectID,
               y = raw_contigs),
           fill = "#282E2C",
           width = 1.15) +
  geom_col(aes(x = CorrectID,
               y = post_qc_absolute_seqs),
           fill = "#299B47",
           width = 1.15) +
  facet_grid(cols = vars(fct_relevel(Species,
                                    c("H. hippopus",
                                      "H. porcellanus",
                                      "T. crocea",
                                      "T. squamosa",
                                      "T. noae",
                                      "T. maxima",
                                      "T. derasa",
                                      "T. gigas"))),
             scales = "free",
             space = "free") +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.line = element_line(),
        axis.text.x = element_blank(),
        panel.spacing = unit(0.1,'lines'),
        axis.title = element_blank(),
        axis.ticks.y =  element_line())
```

Save.

```{r}
A.2_ITS2 |> 
  ggsave(filename = "Sequence stats.svg", 
         dpi = 900,
         width = 16.5, 
         height = 4,
         unit = "cm")
```

Stats.

```{r}
A.1_RawContigsTotal <- 
  A.1_ITS2_2 |> 
  summarise(Total = sum(raw_contigs),
            Mean = mean(raw_contigs),
            SD = sd(raw_contigs))

A.1_PostQCsTotal <- 
  A.1_ITS2_2 |> 
  summarise(Total = sum(post_qc_absolute_seqs),
            Mean = mean(post_qc_absolute_seqs),
            SD = sd(post_qc_absolute_seqs))
```

Save supplementary table.

```{r}
A.1_ITS2_3 <- # for saving
  A.1_ITS2_2 |> 
  select(1:5) |> 
  rename(SampleID = 1,
         Region = 3,
         RawContigs = 4,
         PostQC = 5)

A.1_ITS2_3 |> # save
  write_csv("ST - Sequencing stats.csv")
```

Save supplementary table for ITS2 sequences.

```{r}
A.1_ITS2_4 <- # exact its2 sequences
  A.1_ITS2_2 |> 
  select(1,6:last_col()) |> 
  pivot_longer(!CorrectID,
               names_to = "ITS2Sequence",
               values_to = "Count") |> 
  group_by(ITS2Sequence) |> 
  summarise(Count = sum(Count)) |> 
  ungroup() |> 
  mutate(Clade = str_extract_all(ITS2Sequence, "[A-Z]")) |> 
  arrange(Clade, desc(Count)) |> 
  select(!last_col())

A.1_ITS2_4 |> # save
  write_csv("ST - ITS2 sequences.csv")
```

# ITS2 type profiles

```{r}
B.1_ITS2TypeProf <- 
  read_tsv("../raw_data/20250829T091611/its2_type_profiles/4_first_analysis_20250829T091611.profiles.absolute.abund_and_meta.txt")

B.1_ITS2TypeProf_1 <- # tidy
  B.1_ITS2TypeProf |> 
  slice(-(1:5)) %>% # Remove first 5 rows (unnecessary rows)
  row_to_names(1) |>  # Move first row to column header
  rename(sample_name = 2) |>  # Name 2nd column
  select(-1) |>  # Remove first column (unneeded)
  drop_na() |> 
  type_convert() 

B.1_ITS2TypeProf_2 <- # for saving
  B.1_ITS2TypeProf_1 |>  
  pivot_longer(!sample_name,
               names_to = "ITS2TypeProfile",
               values_to = "Count") |> 
  group_by(ITS2TypeProfile) |> 
  summarize(Count = sum(Count))

B.1_ITS2TypeProf_2 |> # save
  write_csv("ST - ITS2 type profiles.csv")
```
