---
title: "COI compiler"
format: html
theme: 
  light: zephyr
  dark: cyborg
toc: true
page-layout: full
number-sections: true
number-depth: 3
editor: visual
self-contained: true
---

# Notes

A naming convention for dataframes was made.

1.  Objects were named with a prefix per section arranged alphabetically.
2.  A 2nd prefix was added to arrange all the data accordingly:
    -   Dataframes: .**1**
    -   Plots: .**2**
    -   Stats: .**3** (those in dataframe format)
    -   Others:
        -   Vectors: .**vec**
        -   Trees: .**tre**
        -   Distances: .**dis**
        -   Functions: .**fun**
3.  If a dataframe will be used multiple times in case a raw dataframe is needed and an edited one should not supersede the raw one, add a suffix at the end (i.e., **\_1** for first edit, **\_2** for second....).
4.  Limit 3rd prefix names to three letters and if 2 prefixes are needed, separate with "**.**".
5.  If an object is just an auxiliary object, put a **Z** prefix.

# Load pkgs

```{r}
pacman::p_load(readr,
               dplyr,
               stringr,
               Biostrings,
               ape, 
               pegas,
               svglite)
```

# Load data

```{r}
fasta <- read_lines("GCCOISeqs_AlignedAllAddKeep_20251022_clipkitsmartgap.fa")
```

# Tidy

```{r}
df <- tibble(line = fasta) %>%
  mutate(is_header = str_starts(line, ">"),
         group = cumsum(is_header)) %>%
  group_by(group) %>%
  summarise(
    header = str_remove(dplyr::first(line), "^>"),
    sequence = str_c(line[!is_header], collapse = ""),
    .groups = "drop"
  ) %>%
  select(header, sequence)

df
```

# *T. squamosa*

Filter.

```{r}
Tsquamosa <- 
  df |> 
  filter(str_detect(header, "TS"))%>%
  mutate(header = str_extract(header, "TS\\d+")) |> 
  dplyr::rename("SampleID" = header)

TsquamosaMeta <- 
  read_csv("TS - Metadata.csv")
```

Haplotyping

```{r}
# Convert to a dnabin object (for pegas)
TsquamosaDNA <- 
  Tsquamosa |> 
  pull(sequence) |> 
  strsplit(split = "") %>%
  do.call(rbind, .)

rownames(TsquamosaDNA) <- 
  Tsquamosa$SampleID
  
TsquamosaDNA <- 
  as.DNAbin(TsquamosaDNA)
image.DNAbin(TsquamosaDNA) # Check distribution

# Export for re-clipping (if there are gaps)
TsquamosaDNA |> 
  write.FASTA("Tsquamosa alignment.fasta")

# Re-import (TS34 and TS07 were removed due to many gaps)
TsquamosaDNA <- 
  read_lines("Tsquamosa alignment - Clipped.fas")

TsquamosaDNA <- 
  TsquamosaDNA %>%
  tibble(line = .) |> 
  mutate(is_header = str_starts(line, ">"),
         group = cumsum(is_header)) %>%
  group_by(group) %>%
  summarise(
    header = str_remove(dplyr::first(line), "^>"),
    sequence = str_c(line[!is_header], collapse = ""),
    .groups = "drop"
  ) %>%
  select(header, sequence)

TsquamosaDNA <- 
  TsquamosaDNA |> 
  pull(sequence) |> 
  strsplit(split = "") %>%
  do.call(rbind, .)

rownames(TsquamosaDNA) <- 
  ((Tsquamosa$SampleID) |> 
     purrr::discard(~ .x %in% c("TS34", "TS07")))
  
TsquamosaDNA <- 
  as.DNAbin(TsquamosaDNA)
image.DNAbin(TsquamosaDNA) # Check distribution

# Run haplotype
TsquamosaHaps <- 
  haplotype(TsquamosaDNA)
```

Haplotype network.

```{r}
# Tidy
TsquamosaHaps <- sort(TsquamosaHaps, what = "labels")

# Create a data frame linking each individual to its haplotype label
# and to its population from the metadata.
# First, get the haplotype index list.
TsquamosaHapAssign <- 
  attr(TsquamosaHaps, "index")   # a list, each element contains row indices of individuals

names(TsquamosaHapAssign) <-  labels(TsquamosaHaps)


# Build a data frame: one row per individual, with columns "id", "haplotype", "population".
TsquamosaIndividuals <- 
  do.call(rbind, 
          lapply(names(TsquamosaHapAssign), 
                 function(hap) {
                   ids <- rownames(TsquamosaDNA)[TsquamosaHapAssign[[hap]]] # ind names for this haplotype
                   data.frame(SampleID = ids,
                              Haplotype = hap,
                              stringsAsFactors = FALSE)}))

# Merge with metadata to add population information
TsquamosaIndividuals <- 
  TsquamosaIndividuals |> 
  left_join(TsquamosaMeta)

# Export
TsquamosaIndividuals |> 
  write_csv("Tsquamosa - Haplotype - Results.csv")
  
TsquamosaIndividuals <- 
  table(TsquamosaIndividuals$Haplotype, 
        TsquamosaIndividuals$Region)

# Convert to a matrix (pegas expects a matrix)
TsquamosaIndividuals <- as.matrix(TsquamosaIndividuals)

# Plot
TsquamosaHapNet <- 
  haploNet(TsquamosaHaps)

# Open an SVG device
svglite("TsquamosaHaplonet.svg", width = 17/2.54, height = 17/2.54)   # adjust dimensions as needed

# Plot the network with pie charts
plot(TsquamosaHapNet,
     size = attr(TsquamosaHapNet, "freq"), # Scale circles by haplotype frequency
     scale.ratio = 1, # Adjust to avoid circle overlap
     cex = 0.8, # Label size
     pie = TsquamosaIndividuals, # YOUR PIE CHART DATA
    #legend = TRUE, # Display a legend for the pies
     show.mutation = 2
     )

# Close the device – this saves the file
dev.off()
```

## Hd stats

We analysed 428 bp of the mitochondrial COI gene from 24 individuals. A total of 31 polymorphic sites were identified, defining 20 haplotypes (Table 1).

1 Sequence length.

```{r}
# Number of base pairs (columns in the alignment)
seq_length <- ncol(TsquamosaDNA)
print(seq_length)
```

2 Number of polymorphic (segragating) sites

```{r}
# Find positions of segregating sites
poly_sites <- seg.sites(TsquamosaDNA)

# Count them
num_poly <- length(poly_sites)
print(num_poly)
```

3 Number of Haplotypes

```{r}
# Number of distinct haplotypes
num_haplotypes <- nrow(TsquamosaHaps)
print(num_haplotypes)
```

Overall haplotype diversity was Hd = 0.99 ± 0.03, and nucleotide diversity was π = 0.008 ± 0.001. Among populations, Hd ranged from 0.45 (Population C) to 0.92 (Population A) (Table 2).

1 Overall haplotype diversity (Hd)

```{r}
Hd <- 
  hap.div(TsquamosaHaps)

Hd_var <- 
  hap.div(TsquamosaHaps, variance = T)

Hd_sd <- 
  Hd_var[2] |> 
  sqrt()

N <- 
  nuc.div(TsquamosaHaps)

N_var <- 
  nuc.div(TsquamosaHaps, variance = T)

N_sd <- 
  N_var[2] |> 
  sqrt()
```
